\documentclass[../main.tex]{subfiles}
\begin{document}
\chapter{Group Actions}
Groups become groups of symmetries when they ``act''.
\section{Definition and Examples}
\begin{definition}[Group Action]
  An \textit{action} of a group $G$ on a set $X$ is a map $G \times X \to X$, $(g, x) \mapsto gx$ such that:
  \label{actionDef}
  \begin{enumerate}
    \item $ex = x$ for all $x \in X$
    \item $(g \cdot h)x = g(hx)$ for all $g, h \in G$ and $x \in X$
  \end{enumerate}
\end{definition}
We write $G \acts X$.
\begin{example}
  \label{actionExamples}
  \begin{enumerate}
    \item For any group $G$ and set $X$, then $gx = x$ defines the \textit{trivial action}.
    \item $\Sym(X) \acts X$ for any set $X$ by $fx = f(x)$.
    \item If $G \acts X$ and $H \leq G$ then $H \acts X$ by just considering elements in $H$.
    \item In particular, $\Isom(\C) \leq \Sym(\C)$ acts on $\C$.
    \item Similarly, $D_{2n}$ acts on $X_n$ (the regular $n$-gon).
      It also acts on its set of vertices, $\{z \in \C : z^{n} = 1\}$.
    \item Every group $G$ acts \textit{on itself} by $g \gamma = g \cdot \gamma$.
      This is called the \textit{left regular action}.
  \end{enumerate}
\end{example}
\section{Actions and Symmetry}
\begin{theorem}
  An action of a group $G$ on a set $X$ is equivalent to a homomorphism $\phi : G \to \Sym(X)$.
  \label{actionHomTheorem}
\end{theorem}
\begin{proof}
  \begin{proofdirection}{Suppose $G \acts X$}
    Consider $t_g : X \to X$, $x \mapsto gx$ for any $g \in G$.
    Now
    \begin{align*}
      t_{g^{-1}}(t_g(x)) &= t_{g^{-1}}(gx) \\
                         &= g^{-1}(gx) \\
                         &= (g^{-1} \cdot g)x \text{ (by \cref{actionDef}, \textbf{ii})}\\
                         &= ex \\
                         &= x \text{ (by \cref{actionDef}, \textbf{i})}
    \end{align*}
    So $t_{g^{-1}} \circ t_g  = \id_X$.
    Similarly $t_g \circ t_{g^{-1}} = \id_X$ so $t_g$ is invertible and therefore a bijection.
    Thus $t_g$ is a permutation so $t_g \in \Sym(X)$.
    So we may define $\phi: G \to \Sym(X)$, $g \mapsto t_g$.

    We now need to prove that $\phi$ is a homomorphism for $g, h \in G$, $x \in X$.
    \[
      (t_g \circ t_h)(x) = t_g(t_h(x)) = t_g(hx) = g(hx) = (g \cdot h)x = t_{gh}(x)
    \]
    So $t_g \circ t_h = t_{gh}$.
    Therefore $\phi(g \cdot h) = t_{gh} = t_g \circ t_h = \phi(g)\phi(h)$ so $\phi$ is a homomorphism.
  \end{proofdirection}
  \begin{proofdirection}{Suppose we have a homomorphism $\phi: G \to \Sym(X)$}
    We may define an action of $G$ on $X$ by $gx = \phi(g)(x)$.
    We now need to check that this is an action:
    \begin{enumerate}
      \item $ex = \phi(e)(x) = \id_X(x) = x$
      \item $(gh)x = \phi(gh)(x) = (\phi(g) \circ \phi(h))(x) = \phi(g)(\phi(h)(x)) = \phi(g)(hx) = g(hx)$
    \end{enumerate}
    So it satisfies both axioms of \cref{actionDef}.
  \end{proofdirection}
\end{proof}
\begin{remark}[Intuition]
  This means that given a group action of $G$ on $X$ then there is a permutation representation of the action given by $\phi: G \to \Sym(X)$.
  That is, for any $g \in G$ there is a corresponding permutation of $X$, $\phi(g)$, such that $gx = \phi(g)(x)$.

  Similarly if we have a homomorphism $\phi: G \to \Sym X$, there is a corresponding group action of $G$ on $X$ given by $gx = \phi(g)(x)$.

  This correspondence is one-to-one.
\end{remark}
\begin{theorem}[Cayley's Theorem]
  Every group $G$ is isomorphic to a subgroup of some $\Sym(X)$.
  Furthermore, if $G$ is finite then we may choose $X$ to also be finite.
\end{theorem}
\begin{proof}
  Let $X = G$.
  \Cref{actionExamples} \textbf{vi} defines the regular action  of $G$ on $G$, given by $g\gamma = g \cdot \gamma$ for all $g, \gamma \in G$.
  By \cref{actionHomTheorem}, this is equivalent to a homomorphism $\phi: G \to \Sym(G)$.

  Let $H = \im \phi$.
  From \cref{imKerHom}, $H \leq \Sym G$.
  So we may think of $\phi$ as a surjective homomorphism $\phi: G \to H$.

  We now need to prove that $\phi$ is injective.
  By \cref{surjectiveInjectiveProp}, we can do this by proving that $\ker \phi  = \{e\}$.
  Take $g \in \ker \phi$ so $\phi(g) = \id_G$.
  Therefore:
  \[
    g \gamma = \phi(g)(\gamma) = \id_G(\gamma) = \gamma
  \]
  for all $\gamma \in G$.
  In particular, we can pick $\gamma = e$ so $ge = e$.
  But from the definition of the regular action, we also have $g\gamma = g \cdot \gamma$ so $ge = g \cdot e = g$.
  Thus $g = e$ so $\ker \phi = \{e\}$.
  Therefore $\phi$ is injective.

  Hence $\phi$ is bijective and a homomorphism so $G \cong H \leq \Sym(G) = \Sym(X)$  as required.
  Finally, since $G = X$ it is certainly true that $|G| < \infty \implies |X| < \infty$.
\end{proof}
\section{Orbits and Stabilisers}
We can learn a huge amount about groups by thinking about actions.
Orbits and stabilisers can help us do this.
\subsection{Definitions}
\begin{definition}[Orbit]
  If $G \acts X$ and $x \in X$ then the \textit{orbit} of $x$ is the set:
  \[
    Gx = \{y \in X : \exists g \in G \text{ such that } y = gx\}
  \]
\end{definition}
\begin{definition}[Stabiliser]
  If $G \acts X$ and $x \in X$ then the \textit{stabiliser} of $x$ is the set:
  \[
    \Stab_G(x) = \{g \in G : gx = x\}
  \]
\end{definition}
\begin{remark}[Warning]
  Some sources write $G_x$ instead of $\Stab_G(x)$.
\end{remark}
\begin{definition}[Transitive]
  If $Gx = X$, then we say the action is \textit{transitive}.
\end{definition}
\begin{definition}[Faithful]
  If every element $g \in G$ except $e$ has an $x \in X$ such that $gx \neq x$ then the action is \textit{faithful}.
\end{definition}
\begin{remark}
  If $G$ acting on $X$ is faithful if the associated homomorphism $G \to \Sym(X)$ is injective.
\end{remark}
\begin{proposition}
  Suppose $G \acts X$ then:
  \begin{enumerate}
    \item $\Stab_G(x) \leq G$ for any $x \in X$.
    \item The set of all orbits $\{Gx : x \in X\}$ form a \textit{partition} of $X$.
  \end{enumerate}
\end{proposition}
\begin{proof}
  \begin{enumerate}
    \item We need to check that $\Stab_G(x)$ satisfies the definition of a subgroup.
      \begin{itemize}
        \item \textbf{Closure - }
          If $g, h \in \Stab_G(x)$ then:
          \begin{align*}
            (g \cdot h)x &= g(hx) \\
                    &= gx \text{ (Since $h \in \Stab_G(x)$)}\\
                    &= x \text{ (Since $g \in \Stab_G(x)$)}
          \end{align*}
        \item \textbf{Identity - } $ex = x$ by definition of an action so $e \in \Stab_G(x)$.
        \item \textbf{Inverses - }
          For $g \in \Stab_G(x)$,
          \begin{align*}
            g^{-1}x &= g^{-1}(gx) \\
                    &= (g^{-1} \cdot g)x \\
                    &= ex \\
                    &= x
          \end{align*}
          So $g^{-1} \in \Stab_G(x)$.
      \end{itemize}
      Thus $\Stab_G(x) \leq G$.
    \item Similarly to in the proof of \cref{cosetsCover}:

      Notice that $x = ex \in Gx$ so every element is in an orbit, therefore orbits cover $X$.

      If $Gx_1 \cap Gx_2 \neq \emptyset$ then there is some $y = g_1 x_1 = g_2 x_2$ for some $g_1, g_2 \in G$.
      Hence:
      \begin{align*}
        x_1 &= ex_1 \\
            &=(g^{-1}_{1} \cdot g_1) x_1 \\
            &= g^{-1}_1 (g_1 x_1) \\
            &= g^{-1}_1 (g_2 x_2) \text{ (As $g_1 x_1 = g_2 x_2$)} \\
            &= (g^{-1}_{1} \cdot g_2)x_2 \in Gx_2
      \end{align*}
      Now, any $gx_1$ can now be written as $gx_1 = g((g^{-1}_{1} \cdot g_2)x_2) = (g \cdot (g^{-1}_{1} \cdot g_2))x_2 \in Gx_2$.
      Hence $Gx_1 \subseteq Gx_2$.
      Likewise, by relabelling, $Gx_2 \subseteq Gx_1$.
      Thus $Gx_1 = Gx_2$.
      So every element is contained in exactly one orbit so the orbits partition $X$.
  \end{enumerate}
\end{proof}
\begin{remark}[Note]
  By \textbf{ii}, an action is transitive ($Gx = X$) if and only if there is only one orbit, so transitivity is independent of the choice of $x$.
\end{remark}
\begin{example}
  \label{dihedralAction}
  Consider $D_{2n}$ acting on $X_n$, the regular $n$-gon and let $x = 1$.
  Since $D_{2n}$ is a subset of $\Sym(\C)$, it acts by $fx = f(x)$.

  For any $j$, $r^{j}s(x) = r^{j}(x) = e^{\frac{2 \pi i j}{n}} \cdot 1 = e^{\frac{2\pi i j}{n}}$.
  So the action of $r^{j}$ or $r^{j}s$ (i.e. all elements of $D_{2n}$) on $x$ gives $e^{\frac{2 \pi i j}{n}}$.
  Therefore the orbit of $x$, $D_{2n}x = \{e^{\frac{2 \pi i j}{n}} : 0 \leq j < n\}$.

  The above calculation also shows that if $gx = x$ then $j = 0$ so $g$ is either $r^{0} = e$ or $r^{0}s = s$ so $\Stab_{D_{2n}}(x) = \{e, s\}$.
\end{example}
\subsection{Orbit-Stabiliser Theorem}
\begin{theorem}[Orbit-Stabilser Theorem]
  Suppose $G \acts X$ and let $x \in X$.
  The formula:
  \[
    g \Stab_G(x) \mapsto gx
  \]
  defines a well-defined bijection:
  \[
    G / \Stab_G(x) \to Gx
  \]
\end{theorem}
\begin{proof}
  For brevity, let $S = \Stab_G(x)$ and define $\Phi(gS) = gx$.
  We have to check several things about $\Phi$.

  Firstly, we need to check that $\Phi$ is \textit{well-defined}.
  That is, for any $g_1, g_2 \in G$ such that $g_1S = g_2S$ (i.e. they are in the same coset), we need to check that $g_1S$ and $g_2S)$ both get sent to the same thing under the assignment $\Phi$.

  To do this we need to check that $g_1x = g_2x$.
  Now, since $g_1$ and $g_2$ are in the same coset, there is an $s \in S$ such that $g_1 = g_2 \cdot s$.
  Since $s$ is in the stabiliser, $sx = x$ and so:
  \[
    g_1x = (g_2 \cdot s)(x) = g_2(sx) = g_2 x
  \]
  So $\Phi$ is well-defined.

  We now need to check that $\Phi$ is surjective.
  For any $gx \in Gx$,
  \[
    \phi(gS) = gx
  \]
  by definition, so $\Phi$ is surjection.

  We now need to check that $\Phi$ is injective.
  Suppose $\Phi(g_1S) = \Phi(g_2S)$, so $g_1x = g_2x$.
  Now we need to prove that $g_1S = g_2S$.
  Let $s = g^{-1}_{2} \cdot g_1$.
  \begin{align*}
    sx &= (g^{-1}_{2} \cdot g_1)x \\
       &= g^{-1}_{2}(g_1 x) \\
       &= g^{-1}_{2}(g_2 x) \text{ (as $g_1x = g_2x$)}\\
       &= (g^{-1}_{2} \cdot g_2)x \\
       &= ex \\
       &=x
  \end{align*}
  So $s \in S$.
  Therefore:
  \begin{align*}
    g_1 &= (g_2 \cdot g^{-1}_{2}) \cdot g_1 \\
        &= g_2 \cdot (g^{-1}_{2} \cdot g_1) \\
        &= g_2 \cdot s \in g_2S
  \end{align*}
  So $g_1 \in g_2S$ so the cosets $g_1S$ and $g_2S$ overlap but since cosets partition, it follows that $g_1S = g_2S$.
  So $\Phi$ is a well-defined bijection.
\end{proof}
\begin{corollary}
  If $G \acts X$ and $x \in X$ then:
  \[
    |G| = |Gx||\Stab_G(x)|
  \]
\end{corollary}
\begin{proof}
  Orbit-Stabiliser gives a bijection $G/\Stab_G(x) \to Gx$, therefore:
  \[
    |Gx| = |G / \Stab_G(x)|
  \]
  and by Lagrange's theorem (\cref{lagrangesTheorem}):
  \[
    |G| = |G/\Stab_G(x)||\Stab_G(x)|
  \]
  Thus $|G| = |Gx||\Stab_G(x)|$.
\end{proof}
\begin{example}
  Consider again $D_{2n}$ acting on $X_n$ from \cref{dihedralAction}.
  We saw that, if $x = 1$ then $|D_{2n}x| = n$ and $|\Stab_{D_{2n}}(x)| = 2$.
  So by Orbit-Stabiliser, $|D_{2n}| = 2n$.

  This proof is a circular at the moment as we used the group structure to determine the orbits and stabilisers but we could do this without knowing the group structure for an easier proof of the size of $D_{2n}$.
\end{example}
\begin{example}[Symmetries of a Cube]
  Let $G$ be the group of isometries of a cube and let $x$ be the centre of a face.

  The centre of a face can be sent to the centre of any of the other $6$ faces so the orbit of $x$ is the centres of all of the faces, so $|Gx| = 6$.

  To find the stabiliser of $x$, we need to think of isometries that fix the centre of the face, so they must send the face to itself.
  This is just the group of symmetries of a square, $D_{8}$, so $\Stab_G(x) \cong D_8$ and $|\Stab_G(x)| = 8$.

  Therefore $|G| = |Gx||\Stab_G(x)| = 6 \cdot 8 = 48$.
\end{example}
\subsection{Cauchy's Theorem}
The next theorem is a different kind of application of orbit-stabiliser.
\begin{theorem}[Cauchy's Theorem]
  If $|G| < \infty$ and $p$ is a prime that divides $|G|$ then there is a $g \in G$ such that $|g| = p$.
\end{theorem}
\begin{proof}
  Consider the set $X$ of $p$-tuples:
  \[
    X = \{(g_1, \ldots, g_p) : g_i \in G\ \forall i \text{ and } g_1 g_2 \cdots g_p = e\}
  \]
  Define the action of $C_p = \{1, t, t^2, \ldots, t^{p-1}\}$ on $X$ as follows:
  \[
    t^{k}(g_1, \ldots g_p) = (g_{k + 1}, \ldots g_p, g_1, \ldots g_k)
  \]
  This action cycles the elements of the tuple.
  It is easy to see that this action satisfies the two axioms for a group action:
  \begin{itemize}
    \item  $(t^{k} \cdot t^{l})(g_1, \ldots, g_p) = t^{k +l}(g_1, \ldots, g_p)$
    \item $e(g_1, \ldots, g_p) = t^0(g_1, \ldots, g_p) = (g_1, \ldots, g_p)$
  \end{itemize}
  We also need to check that $(g_{k + 1}, \ldots g_p, g_1, \cdots g_k) \in X$.
  Suppose therefore that $g_1 g_2 \cdots g_p = e$.
  For convenience, write:
  \[
    a = g_1 \cdots g_k,\ b = g_{k + 1} \ldots g_p
  \]
  We then have that $a \cdot b = e$, so $b = a^{-1}$ and thus $b \cdot a = e$, that is:
  \[
    g_{k + 1} \cdots g_p g_1 \cdots g_k = e
  \]
  Therefore $(g_{k + 1}, \ldots g_p, g_1, \cdots g_k) \in X$.
  So we have defined a valid action of $C_p$ on $X$.

  We are now going to compute the size of $X$.
  For any choices of $g_1, \ldots, g_{p-1} \in G$ the choice:
  \[
    g_p = g^{-1}_{p-1} \cdots g^{-1}_{1}
  \]
  is the unique choice so that $(g_1, \ldots, g_p) \in X$.
  Hence $|X| = |G|^{p-1}$ as we made a choice between $|G|$ elements $p - 1$ times.

  The action of $C_p$ on $X$ partitions $X$ into orbits:
  \[
    X = C_p x_1 \cup C_p x_2 \cup \cdots \cup C_p x_k
  \]
  By orbit-stabiliser, for each $j$:
  \[
    |C_p x_j| \mid |C_p| = p
  \]
  so since $p$ is prime, either $|C_p x_j| = 1$ or $|C_p x_j| = p$.

  Let $l$ be the number of orbits of size $1$.
  After renumbering, we may assume that:
  \[
    |C_p x_j| = \begin{cases}
    1 & \text{ if } 1 \leq j \leq l \\
    p & \text{ if } j > l
    \end{cases}
  \]
  Since the orbits partition $X$:
  \begin{align*}
    |G|^{p - 1} = |X| &= \sum_{j=1}^{k} |C_p x_j| \\
                      &= \sum_{j=1}^{l} |C_p x_j| + \sum_{j=l+1}^{k} |C_p x_j| \\
                      &= l + p(k - l)
  \end{align*}
  By hypothesis, $p \mid |G|$, so it follows that $p \mid l$.
  Now, note that $x = (g_1, \ldots g_p)$ has an orbit of size $1$ if and only if:
  \[
    g_1 = g_2 = \ldots = g_p
  \]
  as having an orbit of size $1$ means that cycling the elements of the tuple does not affect the tuple.
  In particular, $(e, \ldots, e)$ has an orbit of size 1 so $l > 0$.
  But $p \mid l$ so $l \geq p > 1$ so there has to be at least 1 more orbit of size 1.
  So for some $g \in G$, $g \neq e$:
  \[
    (g, \ldots, g) \in X
  \]
  Therefore by definition of $X$, $g^{p} = e$ so the order of $g$ divides $p$ but $p$ is prime, thus $|g| = p$.
\end{proof}
\section{Conjugation}
\begin{definition}[Conjugate]
  Let $G$ be a group and $g \in G$.
  For any $h \in G$, the element
  \[
    hgh^{-1} \in G
  \]
  is called the \textit{conjugate} of $g$ by $h$.
\end{definition}
\begin{remark}[Intuition]
  The conjugate of $g$ by $h$ has ``the same shape'' as $g$ and is similar to ``changing the coordinates'' of $g$ to assist with a calculation.
\end{remark}
\begin{example}[Conjugate in Abelian Groups]
  If $G$ is an abelian group, then for any $g, h \in G$:
  \[
    h g h^{-1} = g h h^{-1} = g
  \]
  so the only conjugate of $g$ is $g$ itself.
\end{example}
\begin{definition}[Conjugacy Class]
  The \textit{conjugacy class} of an element $g \in G$ is:
  \[
    \ccl(g) = \{h g h^{-1} : h \in G\}
  \]
\end{definition}
\begin{example}
  In Sheet 2 Q6, you will prove that $D_{2n}$ has 1 conjugacy class of reflections if $n$ is odd and 2 if $n$ is even.

  For example, if $n = 6$, there are three axes of reflection that pass through vertices (dashed below) and three axes of reflection that do not pass through vertices (dotted below), so there is two conjugacy classes of reflections that ``look the same''.

  If $n = 5$, there are five axes of reflection that pass through one vertex and one edge, so there is only one conjugacy class.
  \begin{center}
  \begin{tikzpicture}
    \node[draw, minimum size=5cm, regular polygon, regular polygon sides=6] (a) {};
    \foreach \i in {1, 2, 3} {
      \pgfmathsetmacro{\j}{int(\i + 3)}
      \draw[dashed] (a.corner \i) -- (a.corner \j);
    }

    \draw[dotted, thick, gray!70] (a.side 1) -- (a.side 4);
    \draw[dotted, thick, gray!70] (a.side 2) -- (a.side 5);
    \draw[dotted, thick, gray!70] (a.side 3) -- (a.side 6);

    \begin{scope}[xshift=6cm]
      \node[draw, minimum size=5cm, regular polygon, regular polygon sides=5] (b) {};
      \draw[dashed] (b.corner 1) -- (b.side 3);
      \draw[dashed] (b.corner 2) -- (b.side 4);
      \draw[dashed] (b.corner 3) -- (b.side 5);
      \draw[dashed] (b.corner 4) -- (b.side 1);
      \draw[dashed] (b.corner 5) -- (b.side 2);
    \end{scope}
  \end{tikzpicture}
  \end{center}
\end{example}
\begin{proposition}
  A group $G$ acts on itself by conjugation.
  That is
  \[
    g\gamma = g \cdot \gamma \cdot g^{-1}
  \]
  defines an action of $G$ on itself.
\end{proposition}
\begin{proof}
  \begin{itemize}
    \item For any $g \in G$, $ge = g \cdot e \cdot g^{-1} = g \cdot g^{-1} = e$
    \item For any $g, h, \gamma \in G$:
      \begin{align*}
        (g \cdot h)\gamma &= (g \cdot h) \cdot \gamma \cdot (g \cdot h)^{-1} \\
                          &= (g \cdot h) \cdot \gamma \cdot (h^{-1} \cdot g^{-1}) \\
                          &= g \cdot (h \cdot \gamma \cdot h^{-1}) \cdot g^{-1} \\
                          &= g \cdot (h\gamma) \cdot g^{-1} \\
                          &= g(h\gamma)
      \end{align*}
  \end{itemize}
  Therefore it satisfies both axioms of group actions from \cref{actionDef}.
\end{proof}
\begin{remark}[Remarks]
  \begin{itemize}
    \item This is very different to the regular action of $G$ on itself ($g\gamma = g \cdot \gamma$).
    \item $\ccl(g)$ is the orbit of $g$ under this action.
  \end{itemize}
\end{remark}
\begin{definition}[Centraliser]
  The \textit{centraliser} of an element $g \in G$ is defined to be:
  \[
    C_G(g) = \{h \in G: h g h^{-1} = g\}
  \]
\end{definition}
\begin{remark}
  The centraliser of $g$ is the stabiliser of $g$ under the conjugation action.
\end{remark}
\begin{remark}[Note]
  $h g h^{-1} = g \iff hg = gh$, so $C_G(g)$ is the set of elements that \textbf{commute} with $g$.
\end{remark}
\begin{proposition}
  The centraliser of any $g \in G$ is a subgroup of $G$, that is $C_G(g) \leq G$.
  \label{centraliserSubgroup}
\end{proposition}
\begin{proof}
  \begin{itemize}
    \item \textbf{Identity -} $ege^{-1} = g$ so $e \in C_G(g)$.
    \item \textbf{Closure -} If $a, b \in C_G(g)$, that is $aga^{-1} = g$ and $bgb^{-1} = g$, then:
      \begin{align*}
        (ab)g(ab)^{-1} &= (ab)g(b^{-1}a^{-1}) \\
                       &= a (bgb^{-1}) a^{-1} \\
                       &= a g a^{-1} \text{ (As $bgb^{-1} = g$)}\\
                       &= g \text{ (As $aga^{-1} = g$)}
      \end{align*}
      So $ab \in C_G(g)$.
    \item \textbf{Inverses -} If $a \in C_G(g)$ then:
      \begin{align*}
        a^{-1} g (a^{-1})^{-1} &= a^{-1} g a \\
                               &= a^{-1} (a g a^{-1}) a \\
                               &= g \text{ (As $aga^{-1} = g$)}
      \end{align*}
      So $a^{-1} \in C_G(g)$.
  \end{itemize}
  Thus $C_G(g) \leq G$.
\end{proof}
\begin{definition}[Centre of a group]
  The \textit{centre} of a group $G$, $Z(G)$, is defined to be:
  \begin{align*}
    Z(G) &= \{h \in G : hg = gh \text{ for all } g \in G\} \\
         &= \bigcap_{g \in G} C_G(g)
  \end{align*}
  This is exactly the elements of $G$ that commute with everything.
\end{definition}
\begin{remark}
  Since the intersection of two subgroups is also a subgroup (see Example Sheet 1, Q2) and by \cref{centraliserSubgroup} the centraliser of each element is a subgroup, the centre of a group is also a subgroup of $G$.
\end{remark}
\end{document}
