\documentclass[../main.tex]{subfiles}
\begin{document}
\chapter{Cartesian Tensors}
\begin{remark}[Notation]
  In this chapter $\{\vec{e}_1, \vec{e}_2, \vec{e}_3\}$ will refer to any general right handed orthonormal basis for $\R^3$, not just the canonical basis.
  Similarly, we will use $\{\vec{e}_1', \vec{e}_2', \vec{e}_3'\}$ to refer to any other such basis.
\end{remark}
\section{What is a Vector?}
A vector is \textbf{not} just three numbers in a column.
It has existence of its own within a vector space.
So, for example, the vector $\vec{v} = (1, 2, 4)^{\trans}$ just refers to the components of $\vec{v}$ \textbf{with respect to some particular basis}.
If we change the basis, then the components of $\vec{v}$ will change even through but \textbf{the vector $\vec{v}$ itself remains exactly the same}.
\begin{remark}[Nomenclature]
  Changing basis is sometimes referred to as changing the \textit{frame of reference}.
\end{remark}
More precisely, suppose that $\vec{v}$ has components $v_i$ with respect to $\{\vec{e}_i\}$ and $v_i'$ with respect to a new basis $\{\vec{e}_i'\}$, then:
\[
  \vec{v} = v_j \vec{e}_j = v_k' \vec{e}_k'
\]
If we dot both sides with $\vec{e}_i'$, we see that:
\begin{align*}
  v_k' \vec{e}_i' \cdot \vec{e}_k' = v_j \vec{e}_i' \cdot \vec{e}_j \\
  \implies v_k' \delta_{i k} = v_j \vec{e}_i' \cdot \vec{e}_j
\end{align*}
So we can write:
\begin{equation}
  v_i' = R_{i j}v_j \label{vectorTransform}
\end{equation}
where $R = (R_{i j})$ is the matrix with entries given by:
\[
  R_{ij} = \vec{e}_i' \cdot \vec{e}_j
\]
more explicitly:
\[
  R = \begin{pmatrix}
  \vec{e}_1' \cdot \vec{e}_1 & \vec{e}_1' \cdot \vec{e}_2 & \vec{e}'_1 \cdot \vec{e}_3 \\
  \vec{e}_2' \cdot \vec{e}_1 & \vec{e}_2' \cdot \vec{e}_2 & \vec{e}'_2 \cdot \vec{e}_3 \\
  \vec{e}_3' \cdot \vec{e}_1 & \vec{e}_3' \cdot \vec{e}_2 & \vec{e}'_3 \cdot \vec{e}_3 \\
  \end{pmatrix} = \begin{pmatrix}
  {\vec{e}_1'}^{\trans} \\
  \hline
  {\vec{e}_2'}^{\trans} \\
  \hline
  {\vec{e}_3'}^{\trans} \\
  \end{pmatrix}
\]
The $j$-th row of $R$ is just the three components of $\vec{e}_j'$ with respect to the original basis $\{\vec{e}_i\}$.
Since the three rows of $R$ are orthonormal vectors, we deduce that $R$ is an orthogonal matrix so we expect $R R^{\trans} = I$:
\[
  R R^{\trans} = \begin{pmatrix}
  {\vec{e}_1'}^{\trans} \\
  \hline
  {\vec{e}_2'}^{\trans} \\
  \hline
  {\vec{e}_3'}^{\trans} \\
  \end{pmatrix}
  \Biggl(\vec{e}_1' \Bigm| \vec{e}_2' \Bigm| \vec{e}_3'\Biggr) = \begin{pmatrix}
  \vec{e}_1' \cdot \vec{e}_1' & \vec{e}_1' \cdot \vec{e}_2' & \vec{e}'_1 \cdot \vec{e}_3' \\
  \vec{e}_2' \cdot \vec{e}_1' & \vec{e}_2' \cdot \vec{e}_2' & \vec{e}_2' \cdot \vec{e}_3' \\
  \vec{e}_3' \cdot \vec{e}_1' & \vec{e}_3' \cdot \vec{e}_2' & \vec{e}_3' \cdot \vec{e}_3' \\
  \end{pmatrix} = I
\]
as expected.
Since $R$ is orthogonal, this means that $R^{\trans}$ is the inverse of $R$ hence:
\begin{equation}
  R_{i k}(R^{\trans})_{k j} = R_{i k}R_{j k} = \delta_{i j} \text{ and } (R^{\trans})_{i k}R_{k j} = R_{k i}R_{k j} = \delta_{i j} \label{RDelta}
\end{equation}
These are very important identities when working with tensors so it is useful to commit these to memory.

If we expand the determinant of $R$ along the first row, we see that:
\[
  \det R = \det \begin{pmatrix}
  {\vec{e}_1'}^{\trans} \\
  \hline
  {\vec{e}_2'}^{\trans} \\
  \hline
  {\vec{e}_3'}^{\trans} \\
  \end{pmatrix} = \vec{e}_1' \cdot (\vec{e}_2' \times \vec{e}_3') = 1
\]
since $\{\vec{e}_i'\}$ is a right handed orthonormal basis.
So we know that $R$ is orthogonal and $\det R = 1$, thus, as we saw in IA Vectors and Matrices, $R$ must be a rotation matrix.

So $v_i' = R_{i j}v_j$ (\cref{vectorTransform}) for some rotation matrix $R$ defines the transformation rule for vectors.
Any set of three numbers in a column that doesn't obey this rule \textbf{cannot be a vector}.
We can also derive the inverse of this transformation starting from \cref{vectorTransform}:
\begin{align*}
  v_j' = R_{j k}v_k \implies R_{ji}v_j' &= R_{j i}R_{j k}v_k \\
                                        &= (R^{\trans} R)^{i k} v_k \text{ using \cref{RDelta}}\\
                                        &= \delta_{i k}v_k = v_i
\end{align*}
Therefore, the inverse is:
\begin{equation}
  v_i = R_{j i}v_j' \label{invVectorTransform}
\end{equation}
\begin{remark}
  These transformations look just like a map between two different vectors $\vec{v}$ and $\vec{v}'$ where $\vec{v}' = R \vec{v}$ and $\vec{v} = R^{\trans} \vec{v}'$.

  This is a correct interpretation of the equations, however, it is very misleading because in this case $v_i'$ and $v_j$ refer to the components of the \textbf{same vector} $\vec{v}$, just with respect to different bases, not two different vectors.
\end{remark}
\section{What is a Tensor?}
\subsection{Definition}
\begin{definition}[Rank $n$ Tensor]
  A \textit{tensor $T$ of rank $n$} is an array of numbers $T_{ijk\ldots}$ (with respect to a given basis) with $n$ indices which, under a change of basis, transforms according to:
  \begin{equation}
    T_{ijk\ldots}' = R_{i p}R_{j q}R_{k r} \cdots T_{p q r\ldots} \label{tensorTransform}
  \end{equation}
  where $T_{ijk\ldots}'$ are the components of $T$ with respect to the new basis and $R$ is the relevant $3\times 3$ rotation matrix defined by $R_{i j} = \vec{e}_i' \cdot \vec{e}_j$.
\end{definition}
Confusingly, we will often refer to $T_{ijk\cdots}$ as being a tensor, despite it really only being the components of a tensor $T$ with respect to a particular basis.
\begin{remark}
  The rank of a tensor is sometimes referred to its \textit{order}.
  However, rank and dimension are \textbf{not} the same, we will work exclusively in 3 dimensions, so each of the $n$ indices of $T$ can take values in $\{1, 2, 3\}$.
\end{remark}
\begin{remark}
  $T_{ijk\ldots}$ can be visualised as a $3 \times 3 \times \cdots \times 3$ matrix ($n$ times), but this is rarely helpful outside of the cases of $n = 1$ (a 3-vector) or $n = 2$ (a $3\times 3$ matrix).
\end{remark}
The transformation rule for tensors (\cref{tensorTransform}) ensures that $T$ behaves consistently between frames when it interacts with vectors or other tensors.

Similarly to the vector transformation rule, \cref{invVectorTransform}, the tensor transformation rule also has an inverse:
\begin{equation}
  T_{ijk\ldots} = R_{pi}R_{q j}R_{r k} \cdots T_{pqr\ldots}' \label{invTensorTransform}
\end{equation}
\begin{example}[Examples of rank 0 to 2 tensors]
  \begin{enumerate}
    \item A tensor of rank 0 satisfies $T' = T$.
      This means it is a real number that invariant under changes of basis.
      We commonly refer to rank 0 tensors as \textit{scalars}.
    \item A tensor of rank 1 satisfies, according to the transformation rule:
      \[
        T_i' = R_{i p}T_p
      \]
      which is just the transformation rule for vectors (\cref{vectorTransform}).

      So tensors of rank 1 are vectors, since they satisfy the transformation rule for vectors, and any vector is a tensor of rank 1, since they satisfy the transformation rule for rank 1 tensors.
    \item A tensor of rank 2 satisfies:
      \begin{equation}
        T_{i j}' = R_{ip}R_{jq}T_{p q} \label{rank2Transform}
      \end{equation}
      In this case, $T$ can be written as a normal $3 \times 3$ matrix, so the transformation equation becomes:
      \begin{equation}
        T' = RTR^{\trans} \label{matrixTensorLaw}
      \end{equation}
      This is exactly the change of basis formula for matrices (when both sets of bases are orthonormal) that we saw in IA Vectors and Matrices.
  \end{enumerate}
\end{example}
\begin{remark}[Tensors in other coordinate systems]
  \nonexaminable
  In this chapter, we are considering only \textit{Cartesian tensors}.
  It is also possible to define tensors with respect to orthogonal curvilinear coordinate systems (like those in \cref{orthoCurvilinear}), or more general coordinate systems like the space-time of General relativity.
  In this context you may encounter \textit{covariant} vectors $v_i$ (which transform like a change of basis) and \textit{contravariant} vectors $v^{i}$ (which transform oppositely to a change of basis), a distinction which is unnecessary in Cartesians.
\end{remark}
\subsection{Examples of Tensors}
\begin{itemize}
  \item Temperature $\Theta$ is an example of a scalar, it is invariant under a change of basis.
  \item Position $\vec{x}$, velocity $\vec{v}$ and acceleration $\vec{a}$ are all vectors, so are tensors of rank 1.
  \item The Kronecker Delta is a tensor of rank 2. We can actually define the Kronecker delta to be the same for in every basis, that is $\delta_{i j}' = \delta_{i j}$.
    To check it is a tensor, we can just check it satisfies the transformation rule:
    \[
      R_{i p}R_{j q}\delta_{p q} = R_{i p}R_{j p} = (R R^{\trans})_{i j} = \delta_{i j} = \delta_{i j}'
    \]
    so it is a tensor of rank 2.

    We can also write $\delta_{i j}$ as a matrix $T$:
    \[
      RTR^{\trans} = RIR^{\trans} = R R^{\trans} = I = T'
    \]
    since $T = I = T'$.
  \item The Levi-Civita symbol $\levi_{i j k}$ is a tensor of rank 3.
    Again by its definition, it is the same in every basis.
    From IA Vectors and Matrices, we know that:
    \[
      \levi_{i j k}\det A = A_{i p}A_{j q}A_{k r}\levi_{p q r}
    \]
    We can use this fact to check it satisfies the transformation rule:
    \begin{align*}
      R_{i p}R_{j q}R_{k r}\levi_{p q r} &= \levi_{i j k}\det R \\
                                         &= \levi_{i j k} \text{ since $R$ is a rotation} \\
                                         &= \levi_{i j k}'
    \end{align*}
    so it is a tensor of rank 3.
\end{itemize}
\begin{example}[Conductivity Tensor]
  A physical example of a tensor of rank 2 is \textit{conductivity}. \label{conductivityTensor}
  The basic conductivity law is:
  \[
    \vec{J} = \sigma \vec{E}
  \]
  where $\vec{E}$ is the applied electric field and $\vec{J}$ is the resulting electric current.
  In a simple isotropic medium (one which is the same in all directions), $\sigma$ is just a scalar as the conductivity needs to be the same in all directions.
  In an anisotropic medium (one which is not the same in all directions), the relationship between $\vec{J}$ and $\vec{E}$ is still linear but a matrix relation is more suitable.
  For example in a particular basis:
  \[
    \vec{J} = \begin{pmatrix}
    5 & 0 & 0 \\
    0 & 4 & 0 \\
    0 & 0 & 0 \\
    \end{pmatrix}\vec{E}
  \]
  could represent a medium in which the conductivity is high in the $x$-direction but no current can flow in the $z$-direction.
  From this matrix representation, we know that we can write $J_i = \sigma_{i j}E_j$.
  We then call $\sigma$ the \textit{conductivity tensor}.

  Assuming that $\sigma$ is actually a tensor, we will see later in \cref{innerProduct}, that $\vec{J}$ is a rank 1 tensor so transforms properly as a vector.
  Conversely, we will also see later in \cref{quotientTheorem}, that if $\vec{J}$ transforms as a vector, $\sigma$ transforms properly and therefore is a tensor.
  This captures how the tensor transformation rule is exactly what we need to ensure consistency between bases.
\end{example}
\subsection{Non-examples of Tensors}
It is difficult to come up with things that aren't tensors because the concept of the transformation rule is that is obeyed by anything that is mathematically consistent between frames.
\begin{enumerate}
  \item Consider an object $w_i$ of rank 1 given by $\vec{w} = (0, 0, 0)$ with respect to the standard basis and $(1, 0, 0)$ with respect to some other basis.
    Clearly $\vec{w}$ is not a vector as it does not obey the vector transformation rule \cref{vectorTransform}.
    Geometrically speaking, if a vector is the zero vector in one basis, then it must be the zero vector in all bases.
  \item Similarly, a object $W_{i j}$ of rank 2 that is given by $W_{i j} = \delta_{i j}$ in one frame and $-\delta_{i j}$ in another frame is not a tensor.
    We have already shown that:
    \[
      R_{i p}R_{j q}\delta_{p q} = \delta_{i j}
    \]
    and not $- \delta_{i j}$ so the transformation law is not satisfied.
  \item Let $\alpha$ be the first component of a non-zero vector $\vec{v}$.
    $\alpha$ is not a tensor, i.e. not a scalar, because it will be different in different bases.
\end{enumerate}
\section{Sums and Products of Tensors}
\label{tensorSumsAndProducts}
\begin{remark}[Generality of Tensor Proofs]
  It is quite messy to prove results for any general rank of tensor because of the complicated notation.
  In this section, we will instead prove results for particular cases of the rank of the tensor, however the exactly the same methods will work for any other rank, just with extra suffices.
\end{remark}
\subsection{Linear Combinations}
\begin{lemma}[Linear Combination of Tensors]
  A linear combination of two tensors of the same rank is also a tensor of the same rank.
\end{lemma}
\begin{proof}[Rank 2]
  Let $A_{i j}$ and $B_{i j}$ be rank two tensors and let $\alpha$ and $\beta$ be scalars.
  Define $T_{i j} = \alpha A_{i j} + \beta B_{i j}$.
  We can then check that $T$ obeys the tensor transformation rule (\cref{tensorTransform}):
  \begin{align*}
    T_{i j}' &= \alpha' A_{i j}' + \beta' B_{i j}' \\
             &= \alpha A_{i j}' + \beta B_{i j}' \text{ since $\alpha$ and $\beta$ are scalars} \\
             &= \alpha R_{i p} R_{j q} A_{p q} + \beta R_{i p}R_{j q}B_{p q} \text{ since $A$ and $B$ are rank 2 tensors}\\
             &= R_{i p}R_{j q}(\alpha A_{p q} + \beta B_{p q}) \\
             &= R_{i p}R_{j q}T_{p q}
  \end{align*}
  So $T$ is a tensor.
\end{proof}
\subsection{Outer Products}
\begin{definition}[Outer Product]
  The \textit{outer product} or \textit{tensor product} of two tensors $A_{ijk\ldots}$ and $B_{lmn\ldots}$ of ranks $m$ and $n$ respectively is defined as the rank $m + n$ object:
  \[
    T_{ijk\ldots lmn\cdots} = A_{ijk\ldots}B_{lmn\ldots}
  \]
  This relationship is denoted $T = A \otimes B$.
\end{definition}
\begin{lemma}
  The outer product of two tensors $A$ and $B$, that is $T = A \otimes B$, as defined above is also a tensor. \label{outerTensor}
\end{lemma}
\begin{proof}[Rank 2]
  Let $A_{i j}$ and $B_{k l}$ be rank 2 tensors.
  We then just need to check that $T$ satisfies the transformation rule:
  \begin{align*}
    T_{ijkl}' &= A_{i j}' B_{k l}' \\
              &= R_{i p}R_{j q}A_{p q}R_{k r}R_{l s}B_{r s} \text{ since $A$ and $B$ are tensors}\\
              &= R_{i p}R_{j q}R_{k r}R_{l s} A_{j q}B_{r s} \\
              &= R_{i p}R_{j q}R_{k r}R_{l s}T_{pqrs}
  \end{align*}
  so $T$ is a rank 4 tensor.
\end{proof}
\begin{proof}[Rank 1]
  Let $\vec{v}$ and $\vec{w}$ be rank 1 tensors (i.e. vectors).
  We just need to check that $T$ satisfies the transformation rule:
  \begin{align*}
    T_{i j}' &= v_{i}'w_{i}' \\
             &= R_{i p}v_{p}R_{j q}w_{q} \text{ since $\vec{v}$ and $\vec{w}$ are tensors} \\
             &= R_{i p}R_{j q}v_{p}w_{q} \\
             &= R_{i p}R_{j q}T_{p q}
  \end{align*}
  so $T$ is a rank 2 tensor.
\end{proof}
\subsection{Contraction}
\begin{definition}[Contraction]
  A \textit{contraction} of a tensor of rank $n$ occurs when we set two indices equal and sum, resulting in an object of rank $n - 2$.
\end{definition}
\begin{lemma}
  A contraction of a tensor is also a tensor. \label{contractionTensor}
\end{lemma}
\begin{proof}[Rank 2]
  Let $T$ be a tensor of rank 2 and let $\alpha$ be the contraction of $T$, that is, $\alpha = T_{i i}$.
  We then just need to check $\alpha$ is a scalar (i.e. a rank 0 tensor):
  \begin{align*}
    \alpha' = T_{i i}' &= R_{i p}R_{i q}T_{p q} \text{ since $T$ is a tensor} \\
                       &= (R^{\trans} R)_{p q}T_{p q} \text{ using \cref{RDelta}} \\
                       &= \delta_{p q}T_{p q} \\
                       &= T_{p p} = \alpha
  \end{align*}
  So $\alpha$ is a scalar.
\end{proof}
\begin{proof}[Rank 3]
  Let $T$ be a tensor of rank 3 and let $A_{j} = T_{i j i}$ be the contraction of $T$ on its first and third indices.
  \begin{align*}
    A_{j}' &= T_{i j i}' \\
             &= R_{i p}R_{j q}R_{i k} T_{pqk} \text{ since $T$ is a tensor} \\
             &= (R^{\trans} R)_{p k} R_{j q} T_{pqk} \text{ by \cref{RDelta}} \\
             &= \delta_{p k} R_{j q}T_{pqk} \\
             &= R_{j q}T_{pqp} \\
             &= R_{j q}A_{q}
  \end{align*}
  so $A$ is a rank 1 tensor.
\end{proof}
\subsection{Inner Products}
\label{innerProduct}
\begin{definition}[Inner Product]
  An inner product of two tensors $A_{ijk\ldots}$ and $B_{lmn\ldots}$ of ranks $m$ and $n$ respectively occurs when we multiply them, setting one index in each equal and summing, resulting in an object of rank $m + n - 2$.
\end{definition}
\begin{remark}[Nomenclature]
  This is sometimes referred to as ``contracting $A$ and $B$ together''.
\end{remark}
\begin{lemma}
  The inner product of two tensors as defined above is also a tensor.
\end{lemma}
\begin{proof}
  This follows from \cref{outerTensor} and \cref{contractionTensor}.
  We can first do an outer product between the two tensors, resulting in a tensor of rank $m + n$, and then contract on the relevant pair of indices, resulting in a tensor of rank $m + n - 2$.
\end{proof}
\begin{example}[Examples of Inner Products]
  \begin{enumerate}
    \item If $v_i$ and $w_i$ are vectors then $v_iw_i$ is a scalar.
    \item If $A_{i j}$ is a 2nd rank tensor and $v_i$ a vector, then $A_{i j}v_j$ and $A_{i j}v_i$ are both vectors.
    \item If $A_{ijk}$ is a third rank tensor and $B_{i j}$ is a second rank tensor, then $A_{i j k}B_{p j}$ and $A_{i j k}B_{i q}$ are both third rank tensors.
    \item For two vectors $\vec{u}$ and $\vec{v}$, $\vec{u} \cdot \vec{v} = u_iv_i$ must be a scalar because it is an inner product between the rank 1 tensors $\vec{u}$ and $\vec{v}$.

      This means that $\vec{u} \cdot \vec{v}$ does not depend on the basis as it is a rank 0 tensor.
      It therefore makes sense to define the angle between these vectors as:
      \[
        \cos \theta = \frac{\vec{u} \cdot \vec{v}}{\sqrt{(\vec{u} \cdot \vec{u})(\vec{v} \cdot \vec{v})}}
      \]
      which is also a scalar so is invariant under basis changes.
    \item If $\vec{a}$ and $\vec{b}$ are vectors, then $\vec{a} \times \vec{b} = \levi_{i j k}a_jb_k$ is a rank 1 tensor (i.e. a vector) since it is a double inner product of three tensors.
  \end{enumerate}
\end{example}
\section{The Quotient Theorem}
\subsection{Quotient Theorem}
\label{quotientTheorem}
\begin{theorem}[The Quotient Theorem]
  Given an object $T_{ijk\ldots}$ of some rank, if the product of $T$ with an arbitrary tensor $A$ of a given rank is also a tensor, then $T$ is itself a tensor.

  Here, the ``product'', can be an outer product, inner product, or both.
  The important thing is that $A$ is \textbf{arbitrary} apart from its rank.
\end{theorem}
\begin{proof}[Rank 2 \& 1 -- Inner product]
  Let $T$ be an object of rank 2 such that $w_i = T_{i j}v_j$ is a tensor for all vectors $\vec{v}$.

  By assumption $\vec{w}$ is a rank 1 tensor so:
  \begin{align*}
    T_{i j}'v_j' = w_i' &= R_{i p}w_p \text{ using \cref{vectorTransform}} \\
                        &= R_{i p}T_{p q}v_q \\
                        &= R_{i p}T_{p q}R_{j q}v'_j \text{ using \cref{invVectorTransform}} \\
    \implies (T_{i j}' - R_{i p}R_{j q}T_{p q})v'_j &= 0
  \end{align*}
  Since this is true for arbitrary $\vec{v}$, and therefore arbitrary $v'_j$, we have that $T_{i j}' = R_{i p}R_{j q}T_{p q}$ so $T$ obeys the tensor transformation law (\cref{tensorTransform}) and so $T$ is a tensor.
\end{proof}
\begin{remark}
  The above case of the quotient theorem tells us that the conductivity tensor $\sigma$ from \cref{conductivityTensor} must be a tensor since $\vec{J}$ and $\vec{E}$ are vectors and $J_i = \sigma_{i j}E_j$.
\end{remark}
\begin{proof}[Rank 2 \& 3 -- Inner Product]
  Let $T$ be an object of rank 2 such that $W_{i j} = T_{k i}A_{j k}$ is a tensor for all rank 2 tensors $A$.
  \begin{align*}
    T_{k i}' A_{j k}' = W_{i j}' &= R_{i p}R_{j q}W_{p q} \\
                                 &= R_{i p}R_{j q}T_{r p}A_{q r} \\
                                 &= R_{i p}R_{j q}R_{s q}R_{k r}A_{s k}' \text{ using \cref{invTensorTransform}} \\
                                 &= R_{i p}T_{r p}R_{k r}\delta_{j s}A_{s k}' \text{ using \cref{RDelta}} \\
                                 &= R_{k r}R_{i p}T_{r p}A_{j k}' \\
    \implies (T_{k i}' - R_{k r}R_{i p}T_{r p})A_{j k}' &= 0
  \end{align*}
  Since $A_{j k}'$ is arbitrary, $T_{k i}' = R_{k r}R_{i p}T_{r p}$ so $T$ is a tensor.
\end{proof}
\subsection{Tensors as Multi-linear Maps}
\nonexaminable
We have been thinking about tensors in terms of their components.
A more abstract way of thinking about them  is as \textit{multi-linear maps}.
Given a tensor $T_{ijk\ldots}$ of rank $n$, consider the map defined as:
\[
  T: (\R^3)^{n} \to \R,\ T(\underbrace{\vec{a}, \vec{b}, \vec{c}, \ldots}_{n \text{ vectors}}) = T_{ijk\ldots} a_ib_jc_k \cdots
\]
This is multi-linear, that is, it is linear in each of its arguments.
By our definitions so far, since $T_{ijk\ldots}$ is a tensor and $\vec{a}, \vec{b}, \vec{c}, \ldots$ are vectors, $T(\vec{a}, \vec{b}, \vec{c}, \ldots)$ as defined above is an inner product and is therefore a scalar.
Thus this map is well defined since we get the same result in any basis.

For each tensor we can always construct such a map, conversely, given such a map, the multi-linearity shows that it must be of the form above for some coefficients $T_{ijk\ldots}$ which can then be computed in a given basis, for example:
\[
  T_{111\ldots} = T(\vec{e}_1, \ldots, \vec{e}_1)
\]
This is similar to how if we are given a linear map from $\R^3$ to $\R^3$, given a basis, we can create a matrix from it and vice versa.

Furthermore, we can use the Quotient Theorem (\cref{quotientTheorem}) to show that $T_{ijk\ldots}$ is in fact a tensor.
Hence there is an equivalence between tensors and multi-linear maps.

This relation between tensors and maps explains why the transformation rule for rank 2 tensors just looks like the change of basis formula for linear maps from IA Vectors and Matrices.
Recall that in a new basis $A' = P^{\trans} A P$ where $P$ is the orthogonal change of basis matrix.
This is identical to the rank 2 transformation law in \cref{rank2Transform} with $P = R^{\trans}$.

Bu why $P = R^{\trans} = R^{-1}$ rather than just $R$?
This is easiest to think about using the transformation of a vector $\vec{v}$ in 2D.
Suppose that $\{\vec{e}_1', \vec{e}_2'\}$ are the original basis vectors $\{\vec{e}_1, \vec{e}_2\}$ rotated through an angle $\theta$ so that $P$ is a rotation matrix with angle $\theta$.
There are then two ways we could find the components of $\vec{v}$ in the new basis:
\begin{enumerate}
  \item Directly finding the components of $\vec{v}$ with respect to $\{\vec{e}_1', \vec{e}_2'\}$.
  \item Rotate $\vec{v}$ itself by $-\theta$ and then find its components with respect to $\{\vec{e}_1, \vec{e}_2\}$.
\end{enumerate}
So the rank 2 case, \cref{rank2Transform} represents \textbf{(ii)} where $R$ is a rotation by angle $-\theta$, whilst $P$ is the change of basis matrix for \textbf{(i)}.
\section{Symmetries of Tensors}
\subsection{Symmetric and Antisymmetric Tensors}
\begin{lemma}
  If $T_{i j}$ is a tensor, then $T_{j i}$ is also a tensor. \label{tensorIndexSwap}
\end{lemma}
\begin{proof}
  Let $U_{ij} = T_{ji}$, then:
  \begin{align*}
    U_{i j}' &= T_{j i}' \\
             &= R_{j p}R_{i q}T_{p q} \text{ since $T$ is a tensor}\\
             &= R_{i q}R_{j p}U_{q p}
  \end{align*}
  So $T_{j i}$ is a tensor.
\end{proof}
\begin{remark}[Advice]
  Whenever you are asked to prove something is a tensor, it is advisable to define a new object that is exactly what your are trying to show is a tensor as we did above as this prevents confusion with indices and sets a clear goal.
\end{remark}
\begin{definition}
  A tensor $T_{ijk\ldots}$ is \textit{symmetric} in $i$ and $j$ if $T_{ijk\ldots} = T_{jik\ldots}$.
  A tensor can be symmetric in any two two indices, not only adjacent ones.
\end{definition}
\begin{remark}
  For second rank tensors, $T_{i j}$ is symmetric if and only if the corresponding matrix $T = (T_{i j})$ is a symmetric matrix.
\end{remark}
It makes sense to say that the whole tensor is symmetric because we can show that this property is invariant under a change of basis.
For example, for rank 2:
\begin{align*}
  T_{i j}' = R_{i p}R_{j q}T_{p q} &= R_{i p}R_{j q}T_{q p} \text{ using symmetry of $T$}\\
                                   &= R_{j q}R_{i p}T_{q p} \\
                                   &= T_{j i}' \text{ using \cref{tensorTransform}}
\end{align*}
\begin{definition}
  A tensor $T_{ijk\ldots}$ is \textit{anti-symmetric} in $i$ and $j$ if $T_{ijk\ldots} = - T_{jik\ldots}$.
\end{definition}
\begin{remark}[Remarks]
  \begin{itemize}
    \item Similarly to symmetry, anti-symmetric this is also is preserved in all bases.
    \item We call a tensor that is symmetric in every pair of indices \textit{totally symmetric} and we call a tensor that is anti-symmetric in every pair of indices \textit{anti-symmetric}.
  \end{itemize}
\end{remark}
\begin{example}
  $\delta_{i j}$ is a totally symmetric tensor of rank 2 and $\varepsilon_{i j k}$ is a totally anti-symmetric tensor of rank 3.
\end{example}
\begin{lemma}
  If $S_{i j}$ is symmetric and $A_{i j}$ is antisymmetric, then $S_{i j}A_{i j} = 0$. \label{ASProd}
\end{lemma}
\begin{proof}
  \begin{align*}
    S_{i j}A_{i j} &= -S_{j i}A_{j i} \text{ using (anti-)symmetry}\\
                   &= -S_{i j}A_{i j} \text{ swapping dummy indices $i \leftrightarrow j$}\\
    \implies S_{i j}A_{i j} &= 0
  \end{align*}
\end{proof}
The above also applies to higher ranks of tensors.
For example, if $S_{i j k}$ is symmetric in $i$ and $k$ and $A_{pqrs}$ is antisymmetric in $q$ and $r$ then $S_{ijk}A_{pkis} = 0$.
\subsection{Decomposition of 2nd Rank Tensors}
\begin{proposition}[Decomposition into symmetric and antisymmetric parts]
  Any tensor $T_{i j}$ of rank $2$ can be written uniquely in the form: \label{ASDecomp}
  \[
    T_{i j} = S_{i j} + A_{i j}
  \]
  where $S$ is a symmetric tensor and $A$ is an antisymmetric tensor.
\end{proposition}
\begin{proof}\par
  \textbf{Existence}\par
  For existence, let:
  \begin{equation}
    S_{i j} = \frac{1}{2}(T_{i j} + T_{j i}) \text{ and } A_{i j} = \frac{1}{2}(T_{i j} - T_{j i}) \label{tensorDecompStar} \tag{$\star$}
  \end{equation}
  We can check that these satisfy the requirements for symmetry and anti-symmetry:
  \begin{align*}
    S_{j i} &= \frac{1}{2}(T_{j i} + T_{i j}) = \frac{1}{2}(T_{i j} + T_{j i}) = S_{i j} \implies S \text{ is symmetric} \\
    A_{j i} &= \frac{1}{2}(T_{j i} - T_{i j}) = -\frac{1}{2}(T_{i j} - T_{j i}) = -A_{i j} \implies A \text{ is anti-symmetric}
  \end{align*}

  \textbf{Uniqueness}\par
  For uniqueness, suppose that we have $T_{i j} = S_{i j} + A_{i j}$ for some symmetric $S$ and anti-symmetric $A$ then:
  \begin{align*}
    T_{i j} + T_{j i} &= S_{i j} + A_{i j} + S_{j i} + A_{j i} = 2 S_{i j} \\
    T_{i j} - T_{j i} &= S_{i j} + A_{i j} - S_{j i} - A_{j i} = 2 A_{i j}
  \end{align*}
  So $S_{i j}$ and $A_{i j}$ must be as in \cref{tensorDecompStar}.

  This also shows us that $S$ and $A$ must also be tensors since $T$ is a tensor and we can write $A$ and $S$ as a linear combination of the tensors $T_{i j}$ and $T_{ji}$ (which by \cref{tensorIndexSwap}, is also a tensor).
\end{proof}
\begin{remark}
  Note that this decomposition still works for any object of rank 2 (that we do not necessarily know is a tensor), we just then cannot conclude that $S$ and $A$ are tensors.
\end{remark}
\begin{proposition}[Decomposition of antisymmetric second rank tensors]
  If $A_{i j}$ is an antisymmetric tensor, then we can write it in terms of a vector $\vec{\omega}$ as:
  \[
    A_{i j} = \levi_{i j k}\omega_k \text{ where } \omega_i = \frac{1}{2}\levi_{i j k}A_{j k}
  \]
\end{proposition}
\begin{proof}
  $\vec{\omega}$ as defined above is clearly a vector as it is an inner product of a rank 3 tensor $\varepsilon$ and a rank 2 tensor $A$ so will be a tensor of rank $3 + 2 - 2 \cdot 2 = 1$.

  We can then check that $A_{i j}$ is indeed the above expression:
  \begin{align*}
    \levi_{i j k}\omega_k &= \frac{1}{2} \levi_{i j k} \levi_{k l m}A_{l m} \\
                          &= \frac{1}{2}(\delta_{i l}\delta_{j m} - \delta_{i m}\delta_{j l})A_{l m} \\
                          &= \frac{1}{2}(A_{i j} - A_{j i}) \\
                          &= \frac{1}{2}(A_{i j} + A_{i j}) \text{ as $A$ is antisymmetric}\\
                          &= A_{i j}
  \end{align*}
  as required.
\end{proof}
Combining this \cref{ASDecomp}, we can therefore express any second rank tensor $T$ as:
\[
  T_{i j} = S_{i j} + \levi_{i j k}\omega_k
\]
where $S$ is a symmetric tensor and $\vec{\omega}$ is a vector.
\begin{remark}
  The definition of $\vec{\omega}$ is equivalent to the following:
  \[
    A = \begin{pmatrix}
    0 & A_{1 2} & A_{1 3} \\
    -A_{1 2} & 0 & A_{2 3} \\
    -A_{1 3} & -A_{2 3} & 0 \\
    \end{pmatrix} \iff \vec{\omega} = \begin{pmatrix}
    A_{2 3} \\
    -A_{1 3} \\
    A_{1 2} \\
    \end{pmatrix}
  \]
  Since an antisymmetric tensor of rank 2 has only three degrees of freedom, it is not surprising that it can be written in terms of a single vector with three components.
\end{remark}
\subsection{Worked Example}
\begin{remark}[Question]
  Suppose that $C$ is an object of rank 2 such that $C_{i j}S_{i j}$ is a scalar for any \textbf{symmetric} tensor $S$ of rank 2.
  Show that $C$ may not be a tensor but that its symmetric part is.
\end{remark}
We would like to apply the Quotient Theorem but cannot apply it directly as $S$ is \textbf{not arbitrary} since it is restricted to being symmetric.

So we can hope to apply the quotient theorem, let $T$ be an arbitrary second rank tensor.
By \cref{ASDecomp}, we can decompose $T_{i j}$ into
\[
  T_{i j} = S_{i j} + A_{i j}
\]
where $S_{i j}$ is symmetric and $A_{i j}$ is anti-symmetric.
Similarly, we can also decompose $C$ as:
\[
  C_{i j} = C^{(S)}_{i j} + C^{(A)}_{i j}
\]
where $C^{(S)}_{i j}$ is symmetric and $C^{(A)}_{i j}$ is antisymmetric.

Then:
\begin{align*}
  C^{(S)}_{i j} T_{i j} &= C^{(S)}_{i j}(S_{i j} + A_{i j}) \\
                        &= C^{(S)}_{i j}S_{i j} + C^{(S)}_{i j}A_{i j} \\
                        &= C^{(S)}_{i j}S_{i j} \text{ by \cref{ASProd}} \\
                        &= (C_{i j} - C^{(A)}_{i j})S_{i j} \\
                        &= C_{i j}S_{i j} \text{ by \cref{ASProd}}
\end{align*}
Since $S_{i j}$ is symmetric, by assumption, this is a scalar, so $C^{(S)}_{i j}T_{i j}$ is a scalar for any arbitrary rank 2 tensor $T$ and hence $C^{(S)}_{i j}$ is a tensor by the Quotient theorem (\cref{quotientTheorem}).

To find an example in which $C$ is \textbf{not} a tensor, we know that its symmetric part is always a tensor so we need one whose antisymmetric part $C^{(A)}_{i j}$ is not a tensor.
That is, we need to think of with something antisymmetric that is not a tensor.

Consider $C_{i j} = \levi_{1 i j}$ in one frame and $0$ in every other one.
$C_{i j}$ is definitely not a tensor, as it can be $1$ in the original frame and $0$ in another frame.
However, it does satisfy:
\[
  C_{i j}S_{i j} = \begin{cases}
  \levi_{1 i j}S_{i j} & \text{ in the first frame} \\
  0 & \text{ in the others}
  \end{cases}
\]
which, by \cref{ASProd}, in both cases is 0, a scalar.
So this is a valid counterexample.
\begin{remark}
  \nonexaminable
  In fact, $\levi_{1 i j}$ is not a second rank tensor, so for our counter example, we could just have chosen $C_{i j} = \levi_{1 i j}$ in \textbf{all frames}, we would then however have to prove that this is not a tensor.
\end{remark}
\section{Principal Axes}
\subsection{Principal Values and Axes}
For any \textbf{symmetric} second-rank tensor $S$ we can chose a basis in which:
\[
  (S_{i j}) = \begin{pmatrix}
  \lambda_1 & 0 & 0 \\
  0 & \lambda_2 & 0 \\
  0 & 0 & \lambda_3 \\
  \end{pmatrix} \equiv D
\]
This is a direct consequence of the fact from 1A Vectors and Matrices that a real symmetric matrix can be diagonalised using an orthogonal change of basis matrix $P$, satisfying $P^{\trans} S P = D$ for a diagonal matrix $D$ with the eigenvalues of the matrix on its diagonal.
We then let $R = P^{\trans}$ be the corresponding rotation matrix for the tensor transformation law.
We then see that:
\[
  S' = RSR^{\trans} = P^{\trans}SP = D
\]
so $S$ satisfies the matrix form of the rank 2 tensor transformation law in \cref{matrixTensorLaw}, as required.
In general $P$ could have determinant $\pm 1$, however, since swapping two columns in a matrix changes the sign of the determinant, we can always re-order the change of basis so that it is a rotation.

We know from Vectors and Matrices that the $\lambda_i$ along the diagonal the eigenvalues of the matrix.
In this case, they are known as the \textit{principal values} of the tensor, and the \textit{principal axes} are a set of \textbf{orthonormal right-handed axes} parallel to the eigenvectors.

But why should we bother diagonalising a tensor?
We primarily do this because calculations are significantly easier in the frame of the principal axes.
Since we are working with tensors, if are, for example, finding a scalar, then we know the answer will be the same in any frame so it is most convenient to work in the principal axes, where the tensor only has diagonal elements.
\subsection{Inertia Tensor}
For an example of a diagonalisable tensor, we will define the \textit{inertia tensor} for a rigid body $V$.
\begin{remark}[Recap]
  This section uses content from Chapter 5 of IA Dynamics and Relativity.
\end{remark}
Suppose that the body has density $\rho(\vec{x})$ and is rotating with angular velocity $\vec{\omega}$.
The mass in each infinitesimal volume is $\rho(\vec{x})\d{V}$ and since it is rotating with angular velocity $\vec{\omega}$, it will have velocity $\vec{\omega} \times \vec{x}$.

The total angular momentum is then:
\begin{align*}
  \vec{L} &= \int_{V} \rho(\vec{x})(\vec{x} \times (\vec{\omega} \times \vec{x})) \d{V} \\
          &= \int_{V} \rho(\vec{x})((\vec{x} \cdot \vec{x})\vec{\omega} - (\vec{\omega} \cdot \vec{x})\vec{x}) \d{V} \\
  \implies L_i &= \int_{V} \rho(\vec{x}) (x_k x_k \omega_i - \omega_j x_j x_i)\d{V} \\
               &= \int_{V} \rho(\vec{x}) (x_k x_k \delta_{i j} \omega_j - \omega_j x_j x_i)\d{V} \\
               &= I_{i j}\omega_j
\end{align*}
where $I_{i j}$ is the \textit{inertia tensor}, defined as:
\[
  I_{i j} = \int_{V} \rho(\vec{x})(x_k x_k \delta_{i j} - x_ix_j) \d{V}
\]
This tensor is an innate property of the body, independent of the choice of axis in $\vec{\omega}$.
This means that we can then calculate the inertia tensor for the body once and then use it for any choice of $\vec{\omega}$.

Since $\vec{L}$ and $\vec{\omega}$ are vectors, the relationship $L_i = I_{i j}\omega_j$ tells us that $I_{i j}$ must be a tensor by the Quotient Theorem (\cref{quotientTheorem}).
\begin{remark}
  Alternatively, instead of appealing to the Quotient theorem, we know that $I_{i j}$ is a tensor because $x_k x_k \delta_{i j}$ and $x_i x_j$ are tensors (since they are inner/outer products -- \cref{tensorSumsAndProducts}), $\rho$ and $\d{V}$ are scalars unaffected by the rotation, and integral of a tensor is also a tensor.
\end{remark}
From its definition, we see also that it is symmetric by definition and so can be diagonalised.
The above is the inertia tensor about the origin, but we can instead define it about a general point $\vec{a}$ as:
\[
  I_{i j}(\vec{a}) = \int_{V} \rho(\vec{x})(|\vec{x} - \vec{a}|^2 \delta_{i j} - (x_i - a_i)(x_j - x_j)) \d{V}
\]
\begin{remark}[Note]
  In Dynamics and Relativity, the \textit{moment of inertia} about a particular axis is defined as $\int_{V} \rho(\vec{x})r_{\perp}^2 \d{V}$ where $r_{\perp}$ is the perpendicular distance to the axis of rotation.
  When in the standard basis, the components $I_{1 1}, I_{2 2}, I_{3 3}$ of the inertia tensor are just the moments of inertia about the $x$, $y$, and $z$ axes respectively.
\end{remark}
\subsection{Worked Example}
\begin{remark}[Question]
  A cube of uniform density and mass $M$ occupies $V = \{\vec{x} \in \R^3: x, y, z \in [0, a]\}$.
  Find its inertia tensor about the origin and then calculate its principal axes.
\end{remark}
We can calculate the components directly:
\begin{align*}
  I_{1 2} &= \int_{V} \frac{M}{a^3}(x_k x_k \delta_{1 2} - x_1 x_2) \d{V} \\
          &= - \frac{M}{a^3} \int_{0}^{a} \int_{0}^{a} \int_{0}^{a} xy \d{x} \d{y} \d{z} \\
          &= - \frac{1}{4}Ma^2
\end{align*}
and similarly:
\begin{align*}
  I_{1 1} &= \int_{V} \frac{M}{a^3}(x^2 + y^2 + z^2 - x^2) \d{V} \\
          &= \frac{M}{a^3} \int_{0}^{a} \int_{0}^{a} \int_{0}^{a} (y^2 + z^2) \d{x} \d{y} \d{z} \\
          &= \frac{2}{3}Ma^2
\end{align*}
Using the symmetry of the cube and the tensor, the other components follow and so:
\[
  (I_{i j}) = \begin{pmatrix}
  \frac{2}{3} & -\frac{1}{4} & -\frac{1}{4} \\
  -\frac{1}{4} & \frac{2}{3} & -\frac{1}{4} \\
  -\frac{1}{4} & -\frac{1}{4} & \frac{2}{3} \\
  \end{pmatrix}Ma^2
\]
We can easily, but tediously, calculate the eigenvalues to show that the principal values of the tensor are $\frac{1}{6}Ma^2$, $\frac{11}{12}Ma^2$, and $\frac{11}{12}Ma^2$.

It can then be easily shown that the first principal axis is $(\frac{1}{\sqrt{3}}, \frac{1}{\sqrt{3}}, \frac{1}{\sqrt{3}})$.
We also know, again from Vectors and Matrices, that for a symmetric matrix, eigenvectors corresponding to unique eigenvalues are orthogonal.
Since the other two principal values are equal, the corresponding principal axes can any two orthogonal vectors that are orthogonal to the first axis, for example:
\[
  \frac{1}{\sqrt{3}}\begin{pmatrix}
  1 \\
  1 \\
  1 \\
  \end{pmatrix},\ \frac{1}{\sqrt{2}}\begin{pmatrix}
  1 \\
  -1 \\
  0 \\
  \end{pmatrix},\ \frac{1}{\sqrt{6}} \begin{pmatrix}
  1 \\
  1 \\
  -2 \\
  \end{pmatrix}
\]
Note that we quickly find the other two axes by finding one by inspection and then taking the cross product of the first two.
It is also important to check that we have we have ordered and normalised these eigenvectors to be an orthonormal right handed basis, so that they are a valid set of principal axes.

In the basis given by these axis, the inertia tensor is then:
\[
  (I_{i j}) = \begin{pmatrix}
  \frac{1}{6} & 0 & 0 \\
  0 & \frac{11}{12} & 0 \\
  0 & 0 & \frac{11}{12} \\
  \end{pmatrix}Ma^2
\]
Inevitably, the principal axes also coincide with symmetries the physical object.
The first axis is parallel to the line from the origin to the opposite vertex and the other two are any other orthogonal pair of vectors.
\section{Isotropic Tensors}
\begin{definition}[Isotropic Tensor]
  An \textit{isotropic tensor} is one whose \textbf{components are the same in any basis}.
\end{definition}
\begin{remark}[Warning]
  An isotropic tensor is \textbf{not} a tensor which is the same in every basis.
  This is because the tensors stays the same under a change of basis but its \textbf{components change}.
\end{remark}
\subsection{General Forms of Isotropic Tensors}
We have already seen the examples $\delta_{i j}$ and $\levi_{i j k}$.
In fact we can find the most general cases of isotropic tensors for each rank:
\begin{itemize}
  \item Any tensor of rank 0 is isotropic, by definition of scalars.
  \item The only isotropic tensor of rank 1 is the zero-vector as any non-zero vector will have different components in different bases.
  \item The most general isotropic tensor of rank 2 is $\lambda \delta_{i j}$ for some scalar $\lambda$.
  \item The most general isotropic tensor of rank 3 is $\lambda \levi_{i j k}$ for some scalar $\lambda$.
  \item The most general isotropic tensor of rank 4 is:
    \[
      \lambda \delta_{i j}\delta_{k l} + \mu \delta_{i k}\delta_{j l} + \nu\delta_{i l}\delta_{j k}
    \]
    for some scalars $\lambda, \mu, \nu$.
\end{itemize}
\begin{remark}[Exams]
  You are required to remember the general form for ranks 0 -- 3 but are not required to memorise the most general isotropic tensor of rank 4.
  It quite often appears in Tripos exams, but with the form above given.
\end{remark}
\begin{lemma}
  The most general isotropic tensor of rank 2 is $\lambda \delta_{i j}$ for some scalar $\lambda$. \label{general2ndRankIso}
\end{lemma}
\begin{proof}
  Let $T_{i j}$ be any isotropic second rank tensor.
  For ease here, we will use the matrix version of the tensor transformation law from \cref{matrixTensorLaw}, $T' = RTR^{\trans}$.
  We will then chose a basis rotation of $\frac{\pi}{2}$ about the $z$-axis, that is, we take $R$ to be:
  \[
    R = \begin{pmatrix}
    0 & 1 & 0 \\
    -1 & 0 & 0 \\
    0 & 0 & 1 \\
    \end{pmatrix}
  \]
  We can then find $T$ in this basis:
  \begin{align*}
    (T_{i j})' = \begin{pmatrix}
    T_{1 1}' & T_{1 2}' & T_{1 3} \\
    T_{2 1}' & T_{2 2}' & T_{2 3}' \\
    T_{3 1}' & T_{3 2}' & T_{3 3}' \\
    \end{pmatrix} &= \begin{pmatrix}
    0 & 1 & 0 \\
    -1 & 0 & 0 \\
    0 & 0 & 1 \\
    \end{pmatrix} \begin{pmatrix}
    T_{1 1} & T_{1 2} & T_{1 3} \\
    T_{2 1} & T_{2 2} & T_{2 3} \\
    T_{3 1} & T_{3 2} & T_{3 3} \\
    \end{pmatrix} \begin{pmatrix}
    0 & -1 & 0 \\
    1 & 0 & 0 \\
    0 & 0 & 1 \\
    \end{pmatrix} \\
    &= \begin{pmatrix}
    T_{2 2} & -T_{2 1} & T_{2 3} \\
    -T_{1 2} & T_{1 1} & -T_{1 3} \\
    T_{3 2} & -T_{3 1} & T_{3 3} \\
    \end{pmatrix}
  \end{align*}
  By the isotropy of $T$, we must have $T_{i j}' = T_{i j}$ and so:
  \begin{align*}
    T_{1 1} &= T_{2 2} \\
    T_{1 3} &= T_{2 3} = -T_{1 3} \implies T_{1 3} = T_{2 3} = 0 \\
    T_{3 1} &= T_{3 2} = -T_{3 1} \implies T_{3 1} = T_{3 2} = 0
  \end{align*}
  Similarly, carrying out the same process but for rotation of $\frac{\pi}{2}$ about the $y$-axis gives:
  \[
    T_{1 1} = T_{3 3},\ T_{1 2} = T_{3 2} = 0,\ T_{2 1} = T_{2 3} = 0
  \]
  Hence, all off diagonal entries of $T$ vanish and $T_{1 1} = T_{2 2} = T_{3 3} = \lambda$ and so the result follows.
\end{proof}
\begin{remark}
  Interestingly, we only needed to use two very specific rotations to achieve the result.

  If there are symmetries in the problem, it is advisable to pick a rotation that is related to that symmetry to find information about the tensor.
  Furthermore, as we saw above, you often don't need to use all of the facts you are given to arrive at the result and it is much faster to pick a few well chosen bases than to work with a more general rotation.
\end{remark}
There is a similar proof for the general form for third-rank isotropic tensors, however it involves more suffices and we unfortunately cannot use the convenient matrix notation.
\begin{remark}[Physical Significance]
  As an example, consider the conductivity law (from \cref{conductivityTensor}), $J_i = \sigma_{i j}E_j$ in an isotropic medium.
  Since the medium is isotropic, it has the same physical properties in all directions and so $\sigma_{i j}$ must be an isotropic tensor.
  So by \cref{general2ndRankIso}, $\sigma_{i j} = \lambda \delta_{i j}$ for some scalar $\lambda$ and therefore:
  \[
    J_i = \lambda \delta_{i j}E_j = \lambda E_j
  \]
  or equivalently $\vec{J} = \lambda \vec{E}$ and so the direction of $\vec{E}$ does not affect the current.
\end{remark}
\subsection{Isotropic Integrals}
We can use this idea to calculate certain integrals.
This is best illustrated using an example which can then be generalised to other integrals.
\subsubsection{Worked Example}
\begin{remark}[Question]
  Calculate the integral.
  \[
    I_{i j} = \int_{r < a} rx_ix_j \d{V}
  \]
\end{remark}
We observe that $I_{i j}$ is an isotropic second-rank tensor.
Everything in its definition is a tensor and there is no ``preferred direction'' in $\R^3$ so we would expect all sets of axes to produce the same answer.
Although this intuitively makes sense and is sufficient for exams/supervisions, we will provide a formal proof of this in \cref{isotropyProof}.

Therefore, by \cref{general2ndRankIso}, we must have $I_{i j} = \lambda \delta_{i j}$ for some $\lambda$.
We could just calculate the integral for convenient a pair of indices, say $I_{1 1}$, but we can utilise the fact that $\delta_{i i} = 3$ to carry out the following:
\begin{align*}
  3\lambda = I_{i i} &= \int_{r < a} rx_ix_i \d{V} \\
                     &= \int_{r < a} r^3 \d{V} \text{ as $x_ix_i = r^2$} \\
                     &= \int_{r = 0}^{a} \int_{\theta = 0}^{\pi} \int_{\phi = 0}^{2\pi} r^3 \cdot r^2 \sin \theta \d{r} \d{\theta} \d{\phi} \\
                     &= 4\pi \cdot \frac{1}{6}a^6 \\
  \implies \lambda &= \frac{2}{9}\pi a^{6}
\end{align*}
So $I_{i j} = \frac{2}{9}\pi a^{6} \delta_{i j}$.

We can sometimes use a similar trick, even when the starting integral is not isotropic:
\begin{remark}[Extension]
  Consider the vector $\vec{B}$ defined as:
  \[
    \vec{B} = \int_{r < a}  (\vec{x} \cdot \vec{b})|\vec{x}|\vec{x} \d{V}
  \]
  for some non-zero constant vector $\vec{b}$.
\end{remark}
This is clearly not isotropic as it depends on $\vec{b}$ which will have different components in different bases (as it it non-zero).

However, we can write the $i$-th component of $\vec{B}$ as:
\begin{align*}
  \vec{B}_i &= b_j \int_{r < a} x_j r x_i \d{V} \\
            &= b_j \cdot \frac{2}{9}\pi a^{6} \delta_{i j} \\
            &= \frac{2}{9} \pi a^{6} b_i \\
  \implies \vec{B} &= \frac{2}{9} \pi a^6 \vec{b}
\end{align*}
So despite the original integral not being isotropic, we can still use a result about an isotropic integral to help us.
\begin{proposition}[Proof of Isotropy of $I_{i j}$]
  The tensor $I_{i j}$ given by: \label{isotropyProof}
  \[
    I_{i j} = \int_{r < a} rx_ix_j \d{V}
  \]
  is an isotropic second rank tensor.
\end{proposition}
\begin{proof}
  \nonexaminable
  We will first prove that it is a tensor.
  By definition:
  \[
    I_{i j} = \int_{r < a} rx_ix_j \d{x} \d{y} \d{z}
  \]
  Firstly, we note that $r' = r$ as $r^2 = x_ix_i$ is a scalar so will be the same in all bases.
  Transforming into a different basis, we have:
  \begin{align*}
    I_{i j}' &= \int_{r' < a} r'x_i'x_j' \d{x'}\d{y'}\d{z'} \\
             &= \int_{r < a} r R_{i p} x_p R_{j q}x_q |J| \d{x}\d{y}\d{z} \text{ since $r' = r$ and $\vec{x}$ is a vector}
  \end{align*}
  where $J$ is the Jacobian of the transformation $(x, y, z) \mapsto (x', y', z')$.
  However, this is just a rotation so $J = \det R = 1$.
  Therefore:
  \[
    I_{i j}' = R_{i p}R_{j q}\int_{r < a}  rx_p x_q \d{x}\d{y}\d{z} = R_{i p}R_{j q}T_{p q}
  \]
  So it obeys \cref{tensorTransform} and is therefore a tensor.

  Furthermore, to show that it is isotropic:
  \begin{align*}
    I_{i j'} &= \int_{{x'}^2 + {y'}^2 + {z'}^2 < a^2}  \sqrt{{x'}^2 + {y'}^2 + {z'}^2}x_i'x_j'\d{x'}\d{y'}\d{z'} \\
             &= \int_{x^2 + y^2 + z^2 < a^2} \sqrt{x^2 + y^2 + z^2}x_ix_j \d{x}\d{y}\d{z} \\
             &= I_{i j}
  \end{align*}
  We can do this since in the above integral, $x'$, $y'$ and $z'$ are all dummy variables so we can relabel them as $x$, $y$, and $z$, noting also that $x_i'$ and $x_j'$ will be instances of $x'$, $y'$ or $z'$ and so are relabelled to $x_i$ and $x_j$.
\end{proof}
\section{Tensor Calculus}
\subsection{Vector Differential Operators}
When we defined differential operators, such as div, grad, and curl in \cref{chapter1} we did so with respect to the standard basis.
It should not matter if we had used a different basis and so we expect the results to transform correctly.

$\vec{x}$ is a vector so obeys the inverse vector transformation law (\cref{invVectorTransform}), that is:
\begin{align*}
  x_i &= R_{j i}x_j' \\
  \implies \pderiv{x_i}{x_k'} &= R_{j i} \pderiv{x_j'}{x_k'} = R_{j i} \delta_{j k} = R_{k i}
\end{align*}
We can then use this to derive the chain rule for the change of basis.
If we let $\bullet$ be a placeholder for \textbf{any} tensor expression, then:
\begin{equation}
  \pderiv{\,\bullet}{x_i'} = \pderiv{\,\bullet}{x_j} \pderiv{x_j}{x_i'} = R_{i j} \pderiv{\,\bullet}{x_j} \label{chainRuleTensor}
\end{equation}
If we use the notation $\partial_i$ for $\partial/\partial x_i$ then we have:
\[
  \partial_i' = R_{i j}\partial_j
\]
in other words $\nabla$, whose components are $\partial_i$ is a \textit{vector} differential operator as it obeys the vector transformation law in \cref{vectorTransform}.
\begin{lemma}
  If $\vec{F}(\vec{x})$ is a vector field, then $\nabla \cdot \vec{F}$ is a scalar, i.e., it does not depend on the choice of basis.
\end{lemma}
\begin{proof}
  We need to show that $(\nabla \cdot \vec{F})' = \nabla \cdot \vec{F}$:
  \begin{align*}
    (\nabla \cdot \vec{F})' &= \nabla' \cdot \vec{F}' \\
                            &= \pderiv{F_i'}{x_i'} \\
                            &= R_{i j}\pderiv{}{x_j}(R_{i p}F_p) \text{ using \cref{chainRuleTensor}} \\
                            &= (R^{\trans} R)_{j p} \pderiv{F_p}{x_j} \text{ using \cref{RDelta}} \\
                            &= \delta_{j p} \pderiv{F_p}{x_j} \\
                            &= \pderiv{F_j}{x_j} = \nabla \cdot \vec{F}
  \end{align*}
  as required.
\end{proof}
\begin{lemma}
  If $f(\vec{x})$ is a scalar field, then $\nabla f$ is a vector.
\end{lemma}
\begin{proof}
  We need to show that $\nabla f$ obeys \cref{vectorTransform}:
  \begin{align*}
    (\nabla f)_i' &= \nabla'_i f \text{ as $f$ is a scalar} \\
                  &= R_{i j}\pderiv{f}{x_j} \text{ using \cref{chainRuleTensor}} \\
                  &= R_{i j}(\nabla f)_j
  \end{align*}
  as required.
\end{proof}
\begin{lemma}
  If $\vec{F}(\vec{x})$ is a vector field, then $\nabla \times \vec{F}$ is a vector.
\end{lemma}
\begin{proof}
  We need to show that $\nabla \times \vec{F}$ obeys \cref{vectorTransform}:
  \begin{align*}
    (\nabla \times \vec{F})'_i &= (\nabla' \times \vec{F}')_i \\
                               &= \levi_{i j k}'\nabla'_j F_k' \\
                               &= R_{i l}R_{j m}R_{k n}\levi_{l m n} R_{j p} \pderiv{}{x_p} R_{k q}F_{q} \text{ using \cref{chainRuleTensor} and \cref{tensorTransform}} \\
                               &= (R_{j m}R_{j p})(R_{k n}R_{k q})R_{i l} \levi_{l m n} \pderiv{}{x_p}F_q \\
                               &= \delta_{m p}\delta_{n q} R_{i l} \levi_{l m n} \pderiv{}{x_p}F_q \text{ using \cref{RDelta}} \\
                               &= R_{i l}\levi_{l p q} \pderiv{}{x_p}F_q \\
                               &= R_{i l}(\nabla \times \vec{F})_l
  \end{align*}
  as required.
\end{proof}
Similarly to how a vector field is a vector that depends on position, we can generalise this to other ranks of tensors:
\begin{definition}[Tensor Field]
  A \textit{tensor field} is a tensor which depends on position.
\end{definition}
So if $T_{p q r \ldots}(\vec{x})$ is a tensor field of rank $m$ then if we differentiate $n$ times:
\[
  \frac{\partial^{n}}{\underbrace{\partial x_i \partial x_j \partial x_k \cdots}_{n \text{ derivatives}}} T_{p q r \ldots}
\]
then we end up with $m + n$ indices so this is a tensor field of rank $m + n$.
We know that it will be a tensor as we know that $\partial_i$ is a vector differential operator.

For example $\partial T_{i j}/\partial x_k$ is a third-rank tensor field.
\subsection{The Tensor Divergence Theorem}
\nonexaminable
\begin{remark}[Recap]
  Recall the standard divergence theorem (\cref{divergenceTheorem}) for a vector field $\vec{F}$ and volume $V$:
  \[
    \int_{V} \nabla \cdot \vec{F} \d{V} = \int_{S} \vec{F} \cdot \d{\vec{S}}
  \]
  Or equivalently using suffices and summation notation:
  \[
    \int_{V} \pderiv{F_i}{x_i} \d{V} = \int_{\partial V} F_i n_i \d{S}
  \]
\end{remark}
\begin{theorem}[Tensor Divergence Theorem -- Rank 2]
  For a suitable tensor field $T_{i j}(\vec{x})$ in a volume $V$:
  \[
    \int_{V} \pderiv{T_{i j}}{x_i} \d{V} = \int_{\partial V} T_{i j}n_i \d{S}
  \]
\end{theorem}
\begin{proof}
  Let $F_i = a_j T_{i j}$, where $\vec{a}$ is an arbitrary constant vector.
  Since $\vec{F}$ is a inner product of a tensor field $T_{i j}$ and a vector $\vec{a}$, $\vec{F}$ is a vector field and so we can apply the usual divergence theorem:
  \begin{align*}
    \int_{V} \pderiv{F_i}{x_i} \d{V} &= \int_{\partial V} F_i n_i \d{S} \\
    \implies a_j \int_{V} \pderiv{T_{i j}}{x_i} \d{V} &= a_j \int_{\partial V} T_{i j}n_i \d{S}
  \end{align*}
  Since $\vec{a}$ was an arbitrary vector, it follows that:
  \[
    \int_{V} \pderiv{T_{i j}}{x_i} \d{V} = \int_{\partial V} T_{i j}n_i \d{S}
  \]
  as required.
\end{proof}
In fact, we can generalise this to any tensor expression $\bullet$:
\[
  \int_{V} \pderiv{\, \bullet}{x_i} \d{V} = \int_{\partial V} \bullet\; n_i \d{S}
\]
Furthermore, the index $i$ does not need to be the first index in $\bullet$, in fact, $\bullet$ need not have an index $i$ at all.
We can prove this result by introducing as many arbitrary constant vectors as required, namely the rank of $\partial \bullet/\partial x_i$, which matches with the case above where we introduced one arbitrary constant vector.
\end{document}
