\documentclass[../main.tex]{subfiles}
\begin{document}
\chapter{Minimising with Constraints}
\section{Motivation}
Suppose we have a surface $S \subseteq \R^3$ and a function $f: \R^3 \to \R$.
How do we determine the minimum and maximum value of $f$ \textbf{on $S$ specifically}, not just over all of $\R^{3}$.

For example, $S$ might represent a hill in 3D space and $f$ the temperature at that point, then we want to find the hottest and coldest points on the hill.
\section{Setup}
Let $(u, v)$ be a parameterisation of $S$, i.e. $\vec{x} = \vec{x}(u, v)$ on $S$.
We then need to extremise $f(\vec{x}(u, v))$.

At a stationary point $\nabla f = (\pderiv{f}{u}, \pderiv{f}{v}) = \vec{0}$ and so we need to solve:
\begin{align*}
  0 &= \pderiv{}{u}f(\vec{x}(u, v)) = (\nabla f) \cdot \pderiv{\vec{x}}{u} \\
  0 &= \pderiv{}{v}f(\vec{x}(u, v)) = (\nabla f) \cdot \pderiv{\vec{x}}{v}
\end{align*}
But the vectors $\pderiv{\vec{x}}{u}$ and $\pderiv{\vec{x}}{v}$ are linearly independent vectors tangent to $S$ and so since $\nabla f$ is orthogonal to both of them, $\nabla f$ is normal to $S$.

Therefore we have that $\vec{x}_0 \in S$ is a stationary point of $f$ \textbf{on $S$} if and only if $\nabla f(\vec{x}_0)$ is normal to the surface.

However, we often do not have an \textit{explicit} parameterisation of the surface and instead have a surface defined implicitly, for example $g(\vec{x}) = 0$ for some $g: \R^3 \to \R$.
So the problem becomes:
\begin{mini*}
  {\vec{x} \in \R^3}{f(\vec{x})}{}{}
  \addConstraint{g(\vec{x})}{= 0}
\end{mini*}
\begin{remark}[Recap]
  Recall from IA Vector Calculus that $\nabla g$ is normal to the surface $S$.
  This is because if we introduce a parameterisation $h(\vec{x}(u, v)) = 0$ then by taking the partial derivatives with respect to $u$ and $v$ we obtain:
  \[
    (\nabla g) \cdot \pderiv{\vec{x}}{u} = 0 \text{ and } (\nabla g) \cdot \pderiv{\vec{x}}{v} = 0
  \]
\end{remark}
So both $\nabla f$ and $\nabla g$ are normal to $S$ and so a point $\vec{x}_0$ is a stationary point of $f$ if and only if $\nabla f(\vec{x}_0) = \lambda \nabla g(\vec{x}_0)$ for some $\lambda \in \R$.
We can then solve this along with $g(\vec{x}_0) = 0$ to find $\vec{x}_0$.

Both equations are obtained by finding \textit{unconstrained} stationary points $\varphi: \R^4 \to \R$ where:
\[
  \varphi(\vec{x}, \lambda) = f(\vec{x}) - \lambda g(\vec{x})
\]
This is because if we extremis with respect to $\lambda$:
\[
  \pderiv{\varphi}{\lambda} = 0 \iff g = 0
\]
and similarly if we extremis with respect to $\vec{x}$:
\[
  \nabla_{\vec{x}} \varphi = \vec{0} \iff \nabla_{\vec{x}} f - \lambda \nabla_{\vec{x}} g
\]
So minimising $\varphi$ over $\R^4$ is equivalent to finding a stationary point of $f$ in $S$.
We call $\lambda$ a \textit{Lagrange multiplier}, this method can be generalised to $\R^{n}$ in a similar fashion.
\begin{example}
  Consider a box with no lid with side lengths $x, y, z$.
  We want to determine $(x, y, z)$ that minimises the surface area $A$ with  fixed volume $V = L^3/2$.
  Mathematically:
  \begin{mini*}
    {x, y, z}{A(\vec{x}) = 2z(x + y) + xy}{}{}
    \addConstraint{xyz}{= L^3/2}
  \end{mini*}
  \textbf{Direct Method}\par
  We can solve the constraint direction to obtain that $z = \frac{L^3}{xyz} \implies A = \frac{L^3}{x} + \frac{L^3}{y} + xy$.
  We then obtain that a point is stationary if and only if:
  \begin{align*}
    0 &= \pderiv{A}{x} = - \frac{L^3}{x^2} + y \implies x^2 y = L^3 \\
    0 &= \pderiv{A}{y} = - \frac{L^3}{y^2} + x \implies y^2 x = L^3
  \end{align*}
  which has solution $x = y = L$ and $z = L / 2$.
  To check this is minimal we can compute the hessian of $A(x, y)$ and we see that is has positive eigenvalues and so it must be a minimum.

  \textbf{Lagrange Multiplier Method}\par
  Let:
  \[
    \varphi(x, y, z, \lambda) = A(x, y, z) - \lambda\left(xyz - \frac{L^3}{2}\right)
  \]
  To extremise $\varphi$ :
  \begin{align*}
    0 &= \pderiv{\varphi}{z} = 2(x + y) - \lambda xy \implies \lambda = \frac{2(x + y)}{xy} \\
    0 &= \pderiv{\varphi}{x} = 2x + y - \lambda yz \\
      &= 2z + y - \frac{2(x + y)z}{x} \\
      &= \frac{y}{x}(x - 2z) \implies x = 2z
  \end{align*}
  where we can discard the solution with $y = 0$ as this cannot satisfy the constraint.
  \begin{align*}
    0 &= \pderiv{\varphi}{y} = 2z + x - \lambda xz = \frac{x}{y}(y - 2z) \implies y = 2z \\
    0 &= \pderiv{\varphi}{\lambda} = xyz - \frac{L^3}{2}
  \end{align*}
  Upon solving, we therefore obtain $z = L/2, x = y = L$, which is as before.

  This disadvantage of this method is that there is no simple method for checking the nature of the stationary point.
  The Hessian of $\phi$ is always a saddle point because $\pderiv[2]{\varphi}{\lambda}$ and so the Hessian is indefinite.

  \textbf{Conclusion}\par
  Although in this case it was easy to solve the constraint, this is often not the case and so the Lagrange multiplier method is usually preferable.
\end{example}
The Lagrange multiplier method is actually often faster, even when the constraint can be solved.
\begin{example}
  Find the minimum of the quadratic form $f(\vec{x}) = \vec{x}^{\trans} A \vec{x} = x_i A_{i j} x_j$ on $\R^{n}$ on the surface $|\vec{x}|^2 = 1$.
  We define:
  \[
    \varphi(\vec{x}, \lambda) = x_i A_{i j} x_j - \lambda (|\vec{x}|^2 - 1)
  \]
  We see that:
  \begin{align*}
    \pderiv{\varphi}{x_i} &= 0 \iff A_{i j} - \lambda x_i \\
    \pderiv{\varphi}{\lambda} &= 0 \iff |\vec{x}|^2 = 1 \\
  \end{align*}
  This means that the stationary points are the normalised eigenvectors of $A$.
  At a stationary point:
  \[
    f(\vec{x}_0) = x_i A_{i j} x_j = \lambda x_i x_i = \lambda
  \]
  so the eigenvalues of $A$ are the values of $f$ at stationary points and thus the minimum of $f$ is the lowest eigenvalue of $A$.

  \textbf{Alternative Approaches}
  \begin{itemize}
    \item Use a rotation to diagonalise $A$.
    \item Consider: \label{alternativeApproach}
      \[
        \Lambda(\vec{x}) = \frac{f(\vec{x})}{g(\vec{x})}
      \]
      from the constraint we see that $g(\vec{x}) = |\vec{x}|^2$ so:
      \[
        \Lambda(\vec{x}) = \hat{x}_i A_{i j} \hat{x}_j = f(\uvec{x})
      \]
      So the minimum of $f$ with respect to $\uvec{x}$ is equivalent to minimising $\Lambda(\vec{x})$ with respect to $\vec{x} \in \R^n$ (i.e. unconstrained).
      To find the extrema of $\Lambda$:
      \[
        \vec{0} = \nabla \Lambda = \frac{1}{g}\left(\nabla f - \frac{f}{g} \nabla g\right) \iff 0 = \frac{2}{g}\left(A_{i j}x_j - \frac{1}{g}x_i\right) \iff A_{i j}x_j = \Lambda x_i
      \]
      Thus the extrema of $\Lambda$ are the eigenvalues of $A$ with the minimum at the least eigenvalue as before.
  \end{itemize}
\end{example}
\begin{example}
  What probability distribution $\{p_1, \ldots, p_n\}$ satisfying $\sum_{i = 1}^{n} p_i = 1$ maximises the \textit{information entropy} defined as:
  \[
    S = - \sum_{i = 1}^{n} p_i \log_{2}(p_i)
  \]
  We define the function:
  \[
    \varphi(p_1, \ldots, p_n, \lambda)= - \sum_{i = 1}^{n} p_i \log_{2}(p_i) - \lambda \left(\sum_{i = 1}^{n} p_i - 1\right)
  \]
  Recalling that $\log_{2}(p_i) = \frac{\log p_i}{\log 2}$, upon differentiating, we obtain:
  \begin{align*}
    0 &= \pderiv{\varphi}{p_i} = - \log_{2}(p_i) - \frac{1}{\log(2)} - \lambda \\
    0 &= \pderiv{\varphi}{\lambda} = \sum_{i = 1}^{n} p_i - 1
  \end{align*}
  This tells us that all $p_i$ must be equal and so $p_i = \frac{1}{n}\ \forall i \in \{1, \ldots, n\}$.
  So the discrete uniform distribution minimises the information entropy.
\end{example}
If we have multiple constraints and want to find the stationary points of $f: \R^{n} \to \R$ subject to $m < n$ constraints $g_k(\vec{x}) = 0$ for $k \in \{1, \ldots, m\}$, then we can introduce multiple Lagrange multipliers $\lambda_1, \ldots, \lambda_m$ and extremise:
\[
  \phi(x_1, \ldots, x_n, \lambda_1, \ldots, \lambda_m) = f(\vec{x}) - \sum_{k = 1}^{m} \lambda_k g_k(\vec{x})
\]
with respect to $\vec{x}$ and $\lambda_1, \ldots, \lambda_m$.
\end{document}
