\documentclass[../main.tex]{subfiles}
\begin{document}
\chapter{Special Relativity}
\section{Basic Postulates of Relativity}
In 1862, Maxwell's equations predicted the existence of electromagnetic waves that travel in a vacuum at the \textit{speed of light}.
The speed of light $c$ is \textbf{exactly equal to}:
\[
  c = \qty{299792458}{\metre\second^{-1}} \approx \qty{3e8}{\metre\second^{-1}}
\]
\begin{remark}[Note]
  The speed of light is exactly this value because one metre is defined as the distance travelled by light in a vacuum during a time of $1/299792458$ seconds.
\end{remark}

A theory with a ``preferred velocity'' cannot be Galilean invariant.
Galilean theories can only have \textbf{relative velocities}.
In principal, a definite velocity could be acceptable, for example, sound in air also travels at a definite speed $v_{s} \approx \qty{300}{\metre\second^{-1}}$, however, this is \textbf{relative to the air} which is at rest.
So, if you move towards a sound wave with speed $v$, you would measure sound travelling at a speed $v_s + v$.
This lead us to assume that light must also be travelling in some fixed medium, this was called the \textit{ether}.

However, there was an experiments in 1881 by Michelson and Morley that effectively showed that, regardless of the velocity of the observer, you always measure light travelling at the same speed $c$.

This caused a massive crisis in physics, until, in 1905, Einstein postulated that there was no ether and two other postulates:
\begin{enumerate}
  \item The laws of physics are the same in all inertial frames -- This is the same as what Galileo postulated (\cref{galileanRelativity}) and is called the \textit{principal of relativity}.
  \item The speed of light $c$ is the same in \textbf{all inertial frames} -- This is incompatible with inertial frames being related by Galilean transformations as velocities add under Galilean transformations.
\end{enumerate}
\section{1D Lorentz Transformations}
\subsection{Derivation}
We need to change the rules for transforming between frames to ensure that the speed of light is the same in all frames.
We will start with a single spatial dimension $x$ and then build up to higher dimensions later.

Suppose we have frame $S$ with coordinates $(x, t)$ and a frame $S'$ with coordinates $(x', t')$ that moves at a constant velocity $v$ relative to $S$.

According to Galileo, the frames should be related by:
\begin{align*}
  x' &= x - vt \\
  t' &= t
\end{align*}
However, under Einsteins postulates, this is not possible as the speed of light will be different in each frame.

\subsubsection{Finding General Form}
We will start from a general relation:
\begin{align*}
  x' &= f(x, t) \\
  t' &= g(x, t)
\end{align*}
and then show that this leads to a \textit{Lorentz Transformation}.
From the first postulate, the law of inertia is still true, that is, a particle experiencing no forces moves at a constant velocity in all frames.
For such a particle:
\begin{align*}
  \text{In $S$: }& x = A + Bt \\
  \text{In $S'$: }& x' = A' + B't
\end{align*}
Transformations that achieve this are called \textit{affine transformations} which are a linear transformation composed with a translation.
Here, we will choose our frames to share a common origin, i.e. $x = t = 0 \iff x' = t' = 0$ and so $A = A' = 0$.
We can do this as we can always just shift $x$ and $x'$ by constants.
Therefore, the transformation must map all lines passing through the origin in the $(x, t)$-plane to lines passing through the origin in the $(x', t')$-planes, these are precisely linear transformations.

Thus, the transformation must have the form:
\begin{align*}
  x' &= \alpha_1x + \alpha_2t \\
  t' &= \alpha_3x + \alpha_4t
\end{align*}
where $\alpha_i$ are constants that do not depend on $x$ or $t$ but can depend on the relative velocity between frames, since it is constant.

The frame $S'$ is moving at speed $v$ in the frame $S$, but $S'$ is at rest with respect to itself, this means that $x = vt$ should map to $x' = 0$, thus:
\[
  x' = \gamma_v(x - vt)
\]
for some $\gamma_v$ that depends on the relative velocity of the frames $v$.

\subsubsection{Showing $\gamma_v = \gamma_{-v}$}
\begin{lemma}
  Where defined as above, $\gamma_v = \gamma_{-v}$.
\end{lemma}
\begin{proof}
  Consider the frames $\widetilde{S}$ and $\widetilde{S}'$ which are the frames $S$ and $S'$ where the $x$ axis is inverted in direction, that is $\widetilde{x} = -x$ and $\widetilde{x}' = -x'$.

  $S'$ moves at speed $v$ with respect to $S$ and so $\widetilde{S}'$ moves at speed $-v$ with respect to $\widetilde{S}$.
  Therefore:
  \[
    \widetilde{x}' = \gamma_{-v}(\widetilde{x} + \widetilde{t}) = \gamma_{-v}(\widetilde{x} + t)
  \]
  noting that $\widetilde{t} = t$ since time is invariant when inverting just the axis direction.

  Mapping back into the original $x$ axis:
  \[
    -x' = \gamma_{-v}(-x + vt)  \implies x' = \gamma_{-v}(x - vt)
  \]
  and so $\gamma_{v} = \gamma_{-v}$.
\end{proof}
We now define $\gamma \equiv \gamma_{v} = \gamma_{-v}$, noting that it can still depend on velocity.
\begin{remark}
  Another argument for this property is that in 3D, there is no ``preferred'' direction and so $\gamma$ can only depend on the magnitude of the velocity $v^2 = |\vec{v}|^2$ and so $\gamma_v = \gamma_{-v}$.
\end{remark}
\subsubsection{Finding Coefficients in Terms of $\gamma$}
We also require that if we boost by $v$, and then by $-v$, we should get back to the original frame.
That is, a boost by $-v$ is the inverse of a boost by $v$:
\[
  S \xrightarrow{\ v\ } S' \xrightarrow{-v\ }S'' = S
\]
First, considering the boost from $S' \to S''$:
\[
  x'' = \gamma_{-v}(x' + vt') = \gamma(x' + vt')
\]
and then for the boost from $S \to S'$:
\[
  x' = \gamma_{v}(x - vt) = \gamma(x - vt)
\]
Combining the above, we have:
\begin{align*}
  x'' &= \gamma(\gamma(x - vt) + vt') \\
      &= \gamma^2 (x - vt) + \gamma vt'
\end{align*}
We require this to just be $x$ and so:
\[
  \gamma^2 (x - vt) + \gamma vt' = x \implies t' = \gamma t + \frac{1 - \gamma^2}{\gamma v} \cdot x
\]
So we now have $t'$ and $x'$ in terms of $t$ and $x$, with $\gamma$ as the only undetermined quantity.

\subsubsection{Determining $\gamma$ Using the 2nd Postulate}
We can now use the second postulate to determine $\gamma$.
Since the speed of light needs to be the same in both frames, the light ray $x = ct$ must map to $x' = ct'$.
\[
  x' = \gamma(x - vt) = \gamma(ct - vt) = \gamma t(c - v)
\]
and
\[
  t' = \gamma t + \frac{1 - \gamma^2}{\gamma v} \cdot ct
\]
So for $x' = ct'$ to hold, we require:
\[
  \gamma t(c - v) = c\left(\gamma + \frac{1 - \gamma^2}{\gamma} \cdot \frac{c}{v}\right)t
\]
Solving the quadratic for $\gamma$, we have:
\[
  \gamma = \frac{1}{\sqrt{1 - \frac{v^2}{c^2}}}
\]
where we took the positive root to avoid reflecting the $x$ axis.
\begin{remark}
  If $v \geq c$, then we have issues so this transformation only makes sense for $v < c$.
  We will understand this restriction properly later.
\end{remark}
\begin{definition}[Lorentz Transformations]
  \textit{Lorentz Transformations} or \textit{boosts} are transformations of the form:
  \label{lorentzTransforms}
  \begin{align*}
    x' &= \gamma(x - vt) \\
    t' &= \gamma\left(t - \frac{v}{c^2} \cdot x\right)
  \end{align*}
  where $\gamma = \frac{1}{\sqrt{1 - \frac{v^2}{c^2}}}$ is the \textit{Lorentz factor}.
\end{definition}
\begin{remark}[Sanity Check]
  For velocities $v \ll c$, $\gamma \approx 1$ and $\frac{v}{c^2} \to 0$ so we recover the Galilean transformations $x' = x - vt$ and $t' = t$.
  In general, assuming that $v \ll c$ is called taking the ``non-relativistic limit''.
\end{remark}
\subsection{Addition of Velocities}
\label{additionOfVelocities}
Suppose we have a particle moving at a speed $u'$ in a frame $S'$ which is moving at a speed $v$ in frame $S$.
We want to know what speed $u$ the particle moves at in the frame $S$.

To do this, we can use an inverse Lorentz transformation.
By how we derived these transformations, we see that to do the inverse transform, we just need to take $v \mapsto -v$ and so:
\[
  x = \gamma(x' + vt') \text{ and } t = \gamma\left(t' + \frac{v}{c^2}x'\right)
\]
Since the speed of the particle is constant, in $S$ $x = ut$ and in $S'$ $x' = u't'$.
Therefore:
\[
  u = \frac{x}{t} = \frac{x' + vt'}{t' + \frac{v}{c^2}x'} = \frac{\frac{x'}{t'} + v}{1 + \frac{v}{c^2}\frac{x'}{t'}} = \frac{u' + v}{1 + \frac{vu'}{c^2}}
\]
So the relativistic formula for addition of velocities is:
\[
  u = \frac{u' + v}{1 + \frac{vu'}{c^2}}
\]
\begin{remark}[Sanity Check]
  If $v, u' \ll c$, then $\frac{vu'}{c^2} \to 0$ and we just retrieve the standard addition of velocities $u = u' + v$ we would expect for Galilean transformations.
\end{remark}
Furthermore, if $|u'|$ and $|v|$ are both less than $c$, then so is $|u|$.
For example, if $u' = v = \frac{c}{2}$, then Galileo would tell us that $u = c$, whereas we now see that:
\[
  u = c \cdot \frac{\frac{1}{2} + \frac{1}{2}}{1 + \frac{1}{4}} = \frac{4}{5}c
\]
So we cannot make something travel faster than light by moving our frame towards it at a large velocity.

We also see that if $u' = v = c$, then:
\[
  u = \frac{c + c}{1 + c^2/c^2} = c
\]
so if we try to add the speed of light to itself, we get the speed of light back, which is what we wanted.
\subsection{Space-time Diagrams and Simultaneity}
The following type of diagram is called a \textit{space-time diagram}:
\begin{center}
\begin{tikzpicture}[>=stealth, scale=1.5]
  \draw[->] (-0.5, 0)  -- (3, 0) node[right] {$x$};
  \draw[->] (0, -0.5)  -- (0, 3) node[above] {$ct$};

  \node[scale=0.6, above] at (1.2, 3) {world line};
  \node[scale=0.6, above] at (3, 3) {light ray};
  \fill (1, 1) circle (1.5pt) node[right] {$P$} node[left, scale=0.6, xshift=-1] at (1, 1) {event};
  \draw[dashed] (0, 2) -- (2, 0);
  \draw[dashed] (0, 0) -- (3, 3);
  \clip (0, 0) rectangle (3, 3);
  \draw[thick, domain=0:3, smooth, samples=100, rotate=90, yshift=-1cm,xshift=-1] plot (\x,{0.2 *sin(6 *\x r)});
\end{tikzpicture}
\end{center}
\textbf{Features of a space-time diagram}:
\begin{itemize}
  \item A point $P$ on a space-time diagram is called an \textit{event} and represents a point in space and time.
  \item The trajectory of a particle through a space-time diagram is called a \textit{word line} or \textit{worldline}.
  \item The dashed lines on the diagram represent light rays emitted from the particle.
  \item Since the axes are $ct$ and $x$, rays of light must travel at $\pm\ang{45}$ as they have $x = a \pm ct$ for some constant $a$.
\end{itemize}
We can also superpose the axes for a frame $S'$ on the space-time diagram of $S$.
We can use the equations from \cref{lorentzTransforms} to find what the $t'$ and $x'$ axes are mapped to in $S$.

In the frame $S'$, the $ct'$ axis corresponds to the line $x' = 0$:
\[
  x' = 0 = \gamma(x - vt) \implies x = vt \implies ct = \frac{c}{v}x
\]
Similarly, in $S'$, the $x'$ axis corresponds to the line $t' = 0$:
\[
  t' = 0 = \gamma\left(t - \frac{v}{c^2} x\right) \implies x = \frac{c^2}{v}t \implies ct = \frac{v}{c}x
\]
So they are mapped to the lines $ct = \frac{c}{v}x$  and $ct = \frac{v}{c}x$.
Noting that $\frac{c}{v} > 1$ and $\frac{v}{c} < 1$, we can draw this onto the space-time diagram for $S$:
\begin{center}
\begin{tikzpicture}[>=stealth]
  \draw[->] (-3, 0)  -- (3, 0) node[right] {$x$};
  \draw[->] (0, -3)  -- (0, 3) node[above] {$ct$};

  \draw[dashed] (-3, 3) -- (3, -3);
  \draw[dashed] (-3, -3) -- (3, 3);

  \node at (3.3, 1.8) {$x'$};
  \node at (1.9, 3.3) {$ct'$};
  \begin{scope}[decoration={
      markings,
      mark = at position 0.91 with {\arrow{>}}
    }]
    \clip (-3, -3) rectangle (3, 3);
    \draw[rotate=15, postaction={decorate}] (-3, -3) -- (3, 3);
    \draw[rotate=-15, postaction={decorate}] (-3, -3) -- (3, 3);
  \end{scope}
\end{tikzpicture}
\end{center}
\begin{remark}[Note]
  The axes for $S'$ are always symmetric about the line $ct = x$ as $ct = \frac{c}{v}x$ and $ct = \frac{v}{c}x$ are reflections of each other in the line $ct = x$.
\end{remark}
We can also draw lines of constant $t$ and constant $t'$ in the frame $S$.
Lines of constant $t$ are just horizontal lines, whereas lines of constant $t'$ are not horizontal.
Setting $t' = k$, we have:
\[
  k = \gamma\left(t - \frac{v}{c^2}x\right) \implies ct = \frac{v}{c}x + \frac{kc}{\gamma}
\]
So in $S$, the lines of constant $t'$ are all the lines parallel to the $x'$ axis (i.e. $t' = 0$) in $S$.
\begin{center}
\begin{tikzpicture}[>=stealth]
  \node at (0, 3.7) {Lines of constant $t$};
  \draw[->] (-3, 0)  -- (3, 0) node[right] {$x$};
  \draw[->] (0, -3)  -- (0, 3) node[above] {$ct$};

  \fill (-1, 1) circle (1.5pt) node[below] {$P_1$};
  \fill (1, 1) circle (1.5pt) node[below] {$P_2$};

  \foreach {\i} in {-3, -2.5, ..., 3} {
    \draw[thick, dotted] (-3, \i) -- (3, \i);
  }

  \begin{scope}[xshift=7cm]
    \node at (0, 3.7) {Lines of constant $t'$};
    \draw[->] (-3, 0)  -- (3, 0) node[right] {$x$};
    \draw[->] (0, -3)  -- (0, 3) node[above] {$ct$};

    \fill (-1, 1) circle (1.5pt) node[below] {$P_1$};
    \fill (1, 1) circle (1.5pt) node[below] {$P_2$};

    \node at (3.3, 1.8) {$x'$};
    \node at (1.9, 3.3) {$ct'$};
    \begin{scope}[decoration={
        markings,
        mark = at position 0.91 with {\arrow{>}}
      }]
      \clip (-3, -3) rectangle (3, 3);
      \foreach {\i} in {-4, -3.5, ..., 4} {
        \draw[thick, dotted, rotate=30] (-4, \i) -- (4, \i);
      }
      \draw[rotate=15, postaction={decorate}] (-3, -3) -- (3, 3);
      \draw[rotate=-15, postaction={decorate}] (-3, -3) -- (3, 3);
    \end{scope}
  \end{scope}
\end{tikzpicture}
\end{center}
We say that two events are \textit{simultaneous} in a frame if they happen at the same time in that frame.
So $P_1$ and $P_2$ are simultaneous in $S$ as they lie on the same line of constant $t$ but they are not simultaneous in $S'$ as they do not lie on the same line of constant $t'$.

\begin{example}[Einstein's Relativity Train]
  Suppose we have a train travelling along a track at a constant speed with a source of light at the centre of the train that emits light in both directions.

  In the frame of the train, the train has velocity $0$ and so the light rays will reach the front and the back of the train at the same time.
  In the frame of the platform, the train has velocity $v$ so, provided this is non-zero, since the speed of light is the same in both frames, train will have moved forward in the time it takes for the light rays to travel and so they will reach the back of the train first first.

  In Galilean physics, the observer on the platform would see light going at $c + v$ in one direction and $c - v$ in the other, so would also see them reaching both ends at the same time.
\end{example}
We might worry that if different frames can see things at different times, then might we be able to see the cause of an event after the effect in some frame?
Thankfully this is not the case because lines of simultaneity in $S'$ (i.e. constant $t'$) can be at most at $\ang{45}$ in a space-time diagram for $S$.

This is effect can be illustrated by \textit{light cones}:
\begin{center}
\begin{tikzpicture}[>=stealth, scale=1.5]
  \begin{scope}[pattern=north east lines, pattern color=gray!70]
    \fill (0, 3) -- (3, 3) -- (1.5, 1.5);
    \fill (0, 0) -- (3, 0) -- (1.5, 1.5);
  \end{scope}
  \node[scale=0.6, above] at (3, 3) {light ray};
  \draw[->] (0, 0)  -- (3, 0) node[right] {$x$};
  \draw[->] (0, 0)  -- (0, 3) node[above] {$ct$};

  \fill (1.5, 1.5) circle (1pt) node[right] {$P$};
  \fill (1.5, 2.2) circle (1pt) node[right] {$Q$};
  \fill (2.7, 2.2) circle (1pt) node[right] {$R$};
  \draw[dotted, thick] (0, 0.625) -- (3, 2.375);
  \draw[dotted] (0, 1.5) -- (3, 1.5);
  \draw (1.5, 1.5) ++ (0:0.5) arc (0:30:0.5) node[midway, right, scale=0.7] {$< \ang{45}$};
  \draw[dashed] (0, 3) -- (3, 0);
  \draw[dashed] (0, 0) -- (3, 3);

  \node[fill=white, scale=0.5] at (1.5, 0.5) {Past light cone};
  \node[fill=white, scale=0.5] at (1.5, 2.5) {Future light cone};
\end{tikzpicture}
\end{center}
The future light cone contains every event that can be influenced by $P$ and the past light cone contains every event that can influence $P$.

The lines of simultaneity in $S'$ have gradient $\frac{v}{c}$ in the $S$ space-time diagram and so can make at most an angle of \ang{45} with the $x$ axis as $v \to c$.
We can find a frame where $P$ and $R$ (or any other event outside the light cones) are simultaneous since the line joining them makes an angle $<\ang{45}$ and so this could be a line of simultaneity in a different frame.
However, we cannot make $P$ and $Q$ (or any other event in the past or future light cone) simultaneous since this would need a vertical line which cannot be a line of simultaneity in a different frame.

So any event in the future light cone for $P$ must be to the future of $P$ in \textbf{every frame}, and similarly, every event in the past light cone for $P$, must be to the future of $P$ in \textbf{every frame}, this ensures causality.
\begin{remark}
  We can also see this by considering the light rays that bound the light cones.
  By construction, we ensured that light rays map to light rays and so the set of all possible events in the light cones for $P$ are the same in all frames.
\end{remark}
\subsection{Time Dilation}
Suppose we have a clock at rest in the frame $S'$ that ticks at time intervals $T'$.
That is, in frame $S'$, the ticks occur at space-time coordinates $(0, 0)$, $(cT', 0)$, $(2cT', 0)$ and so on.
We want to know when these ticks occur in the frame $S$.

We can use the inverse Lorentz transformation:
\[
  t = \gamma\left(t' + \frac{v}{x^2}x'\right)
\]
Since $x' = 0$ for all ticks in $S'$, the ticks occur at $t = 0,\ \gamma T',\ 2\gamma T'$ and so on, that is, at intervals $T = \gamma T'$.
So if $S'$ is moving at $v \neq 0$ in $S$, then $\gamma > 1$ and so $T > T'$.
That is, for an observer in $S$, there is a longer interval between the ticks of the clock.
\begin{example}[Twin Paradox]
  Suppose we have two identical twins $A$ and $B$.
  Twin $A$ stays on Earth and twin $B$ travels to Neptune and back at almost light speed.
  The twin paradox concerns the question of which twin will be older when $B$ returns.

  In the frame of the Earth, $A$ observes $B$'s clock ticking more slowly and therefore expects $B$ to be younger then they return.
  However, we could also argue that in $B$'s frame, the Earth is moving away from them and so $A$'s clock should be ticking more slowly and therefore $B$ expects $A$ to be younger when they return.

  However, this paradox breaks down due to the asymmetry that twin $B$ has to turn around at some point to return.
  Suppose that from the perspective of $A$, it takes time $T$ for $B$ to arrive at Neptune.
  Since $B$ travels at the same speed back, the full trip takes a time $2T$ from the perspective of $A$:
  \begin{center}
  \begin{tikzpicture}[>=stealth, scale=1.5]
    \draw[->] (0, 0)  -- (3, 0) node[right] {$x$};
    \draw[->] (0, 0)  -- (0, 3) node[above] {$ct$};

    \node[below, scale=0.6] at (0, 0) {Earth};
    \draw[dashed] (1, 0) node[below, scale=0.6] {Neptune} -- (1, 3);
    \draw[dotted] (0, 1.5) -- (1, 1.5);

    \draw[|-|] (-0.1, 0) -- (-0.1, 1.5) node[midway, left] {$cT$};
    \draw[|-|] (-0.1, 1.5) -- (-0.1, 3) node[midway, left] {$cT$};
    \fill (1, 1.5) circle (1.5pt) node[right] {$P$};
    \draw[postaction={decorate}, decoration={
        markings,
        mark = at position 0.25 with {\arrow{>}},
        mark = at position 0.75 with {\arrow{>}}
      }] (0, 0) -- (1, 1.5) -- (0, 3);
    \begin{scope}[xshift=4cm]
      \draw[->] (0, 0)  -- (3, 0) node[right] {$x$};
      \draw[->] (0, 0)  -- (0, 3) node[above] {$ct$};

      \draw[->] (0, 0) -- (3, 2) node[above right] {$x_{\text{out}}'$};

      \draw[dotted] (0, 0.83) -- (3, 2.83) node[scale=0.5, right] {Outward Simultaneity};
      \draw[dotted] (0, 2.167) -- (3, 0.167) node[scale=0.5, right] {Return Simultaneity};

      \node[below, scale=0.6] at (0, 0) {Earth};
      \draw[dashed] (1, 0) node[below, scale=0.6] {Neptune} -- (1, 3);

      \draw[|-|] (-0.1, 0) -- (-0.1, 0.83) node[midway, left] {$c\hat{T}$};
      \draw[|-|, dotted] (-0.1, 0.83) -- (-0.1, 2.167) node[midway, left, scale=0.5] {Missed ticks};
      \draw[|-|] (-0.1, 2.167) -- (-0.1, 3) node[midway, left] {$c\hat{T}$};
      \fill (1, 1.5) circle (1.5pt) node[right] {$P$};
      \draw[postaction={decorate}, decoration={
          markings,
          mark = at position 0.25 with {\arrow{>}},
          mark = at position 0.75 with {\arrow{>}}
        }] (0, 0) -- (1, 1.5) -- (0, 3);
    \end{scope}
  \end{tikzpicture}
  \end{center}
  As discussed above, we then have $2T' = \frac{2T}{\gamma}$ with $\gamma > 1$ and so from the point of view of twin $A$, less time has passed on twin $B$'s clock.

  If we now consider twin $B$'s perspective, they will also see $A$'s clock running slower.
  Similarly $2\hat{T} = \frac{2T'}{\gamma}$, i.e. $A$'s clock appears slower to $B$, however, because the simultaneity in $B$'s frame is different, they see \textbf{$A$'s clock running slower for shorter period of time}.
  This is because $B$ misses some of the ticks of $A$'s clock as there are no lines of simultaneity for $B$ passing through $P$ in either the inward or outward direction that also pass through the ``missing ticks'' region.
  If we account for this jump, then both $A$ and $B$ will agree that $B$ is younger.
\end{example}
\begin{remark}[Example Sheet]
  See Q4 on Example Sheet 4 for a mathematical verification of this.
\end{remark}
\subsection{Length Contraction}
\label{lengthContraction}
Suppose we have a rod with length $L'$ at rest in a frame $S'$.
We define the length of the rod in a frame to be the distance between its endpoints at a fixed time.
\begin{center}
\begin{tikzpicture}[>=stealth]
  \draw[->] (0, 0)  -- (3, 0) node[right] {$x'$};
  \draw[->] (0, 0)  -- (0, 3) node[above] {$ct'$};

  \fill (0, 0) circle (1.5pt) node[below] {$P_1$};
  \fill (1, 0) circle (1.5pt) node[below] {$P_2$};
  \draw[|<->|] (0, 0.1) -- (1, 0.1) node[midway, above] {$L'$};
  \draw (1, 0) -- (1, 3);
  \clip (0, 0) rectangle (3, 3);
\end{tikzpicture}
\end{center}
Now consider a frame $S$ in which the frame $S'$ is moving with velocity $v$:
\begin{center}
\begin{tikzpicture}[>=stealth]
  \draw[->] (0, 0)  -- (3, 0) node[right] {$x$};
  \draw[->] (0, 0)  -- (0, 3) node[above] {$ct$};

  \fill (0, 0) circle (1.5pt) node[below] {$P_1$};
  \fill (1, 0) circle (1.5pt) node[below] {$P_3$};
  \draw[|<->|] (0, 0.1) -- (1, 0.1) node[midway, above] {$L$};
  \draw (1, 0) -- (3, 3);
  \draw (0, 0) -- (2, 3);
  \draw[dashed, ->] (0, 0) -- (3, 2) node[above right] {$x'$};
  \fill (1.8, 1.2) circle (1.5pt) node[below right] {$P_2$};

  \draw[|<->|] (0, 0) -- (1.8, 1.2) node[midway, above] {$L'$};
\end{tikzpicture}
\end{center}
Our objective is to calculate $L$ given $L'$.
We know that:
\begin{align*}
  P_1 = (0, 1) \text{ in both $S$ and $S'$} \\
  P_2 = (ct', x') = (0, L') \text{ in $S'$}
\end{align*}
We can then apply an inverse Lorentz transformation to find $P_2$ in $S$:
\[
  P_2 = (ct, x) = \left(c\gamma\left(0 + \frac{v}{c^2}L'\right), \gamma(L' + v \cdot 0)\right) = \left(\gamma \frac{v}{c} L', \gamma L'\right)
\]
In $S$, the front of the rod at $P_2$ moved with a speed $v$ for a time $t$, so to get $P_3$ in $S$, we have:
\[
  \at{x}{P_3} = \at{x}{P_2} - v \cdot \at{t}{P_2} = \gamma L - v \cdot \frac{\gamma v L'}{c^2} = \gamma L' \left(1 - \frac{v^2}{c^2}\right) = \frac{L'}{\gamma}
\]
Since $P_1$ and $P_3$ are simultaneous in $S$, the length of the rod $L$ in $S$ is $L = \at{x}{P_3} = \frac{L'}{\gamma}$
$v < c$ and so $\gamma > 1$, thus a moving rod appears shorter.
\begin{example}[Ladder and Barn]
  Suppose we have a ladder with length $2L$ and barn of length $L$ open at both ends.
  Person $A$ is holding a ladder and is moving towards the front of the barn and person $B$ is watching from outside the barn.

  In $B$'s frame, the barn is stationary and the ladder is moving.
  If $A$ travels sufficiently fast, then $B$ will see the ladder length contract enough to be short enough to fit entirely inside the barn at some point.

  However, in $A$'s frame, the ladder is stationary and the barn is moving and so the barn will be length contracted and thus the ladder can never fit entirely inside the barn.

  This might at first seem contradictory but can be explained using a space-time diagram:
  \begin{center}
  \begin{tikzpicture}[>=stealth, scale=1.3]
    \draw[->] (0, 0)  -- (3, 0) node[right] {$x$};
    \draw[->] (0, 0)  -- (0, 3) node[above] {$ct$};

    \draw (1, 0) -- (3, 3);
    \draw (0, 0) -- (2, 3);
    \draw[dotted, ->] (0, 0) -- (3, 2) node[right] {$x'$};
    \draw[dotted] (0, 0.33) -- (3, 2.33) node[right, scale=0.5] {Simultaneity in $S'$};

    \draw[dashed, thick] (0.4, 0.6) -- (2.2, 1.8);

    \fill (0.4, 0.6) circle (1pt) (2.2, 1.8) circle (1pt) (1.2, 1.8) circle (1pt);


    \draw (1.2, 0) node[below, scale=0.5] {Front} -- (1.2, 3);
    \draw (2.2, 0) node[below, scale=0.5] {Back} -- (2.2, 3);

    \draw[thick] (1.2, 1.8) -- (2.2, 1.8);
    \draw[dotted] (0, 1.8) node[left, scale=0.5] {Simultaneity in $S$} -- (3, 1.8);
  \end{tikzpicture}
  \end{center}
  Consider the event of the front of the ladder just reaching the back side of the barn.
  In $B$'s frame $S$, this event is simultaneous with the event of rear of the ladder entering the barn and so the ladder is entirely inside the barn for $B$.
  However, if we draw a line of constant $t'$ through this event, we see that in $S'$, it is simultaneous with the rear of the ladder still sticking out of the barn, so from $A$'s perspective, there is never a point where the ladder is fully inside the barn.
\end{example}
\subsection{The Invariant Interval}
We have seen that two frames do not necessarily agree on when and where something happens, however there is one quantity that all frames will always agree on.
Consider two events $P_1$ and $P_2$ with coordinates $(ct_1, x_1)$ and $(ct_2, x_2)$ respectively in the frame $S$ and let $\Delta t = t_2  - t_1$ and $\Delta x = x_2 - x_1$.
\begin{definition}[Invariant Interval]
  The \textit{invariant interval} $(\Delta s)^2$ is:
  \[
    (\Delta s)^2 \equiv c^2 (\Delta t)^2 - (\Delta x)^2
  \]
\end{definition}
We check it is invariant by calculating this in the $S'$ frame.
Since the Lorentz transformations are linear, we can introduce a $\Delta$ to each variable:
\begin{align*}
  \Delta x' &= \gamma(\Delta x - v \Delta t) \\
  \Delta t' &= \gamma\left(\Delta t - \frac{v}{c^2} \Delta x\right)
\end{align*}
Therefore:
\begin{align*}
  c^2(\Delta t')^2 - (\Delta x')^2 &= \gamma^2 c^2 \left(\Delta t - \frac{v}{c^2} \Delta x\right)^2 - \gamma^2 \left(\Delta x - v \Delta t\right)^2 \\
                                   &= (\Delta t)^2(\gamma^2 c^2 - \gamma^2 v^2) + (\Delta x)^2\left(\gamma^2 \cdot \frac{c^2 v^2}{c^4} - \gamma^2\right) \\
                                   &= (\Delta t)^2\left(\frac{c^2}{1 - \frac{v^2}{c^2}} - \frac{v^2}{1 - \frac{v^2}{c^2}}\right) + (\Delta x)^2\left(\frac{1}{1 - \frac{v^2}{c^2}}\right)\left(\frac{v^2}{c^2} - 1\right)\\
                                   &= c^2(\Delta t)^2 - (\Delta x)^2
\end{align*}
So the invariant interval is the same in both frames.

We can also express the invariant interval using matrices as:
\[
  (\Delta s)^2 = \begin{pmatrix}
  c \Delta t & \Delta x \\
  \end{pmatrix} \begin{pmatrix}
  1 & 0 \\
  0 & -1 \\
  \end{pmatrix} \begin{pmatrix}
  c \Delta t \\
  \Delta x \\
  \end{pmatrix}
\]
Since it is linear, we can also write the Lorentz transformation as a matrix:
\[
  \begin{pmatrix}
  c t' \\
  x' \\
  \end{pmatrix} = \begin{pmatrix}
  \gamma & - \frac{\gamma v}{c} \\
  - \frac{\gamma v}{c} & \gamma \\
  \end{pmatrix} \begin{pmatrix}
  ct \\
  x \\
  \end{pmatrix}
\]
Invariance of $(\Delta s)^2$ is equivalent to:
\begin{align*}
  (\Delta s)^2 &= \begin{pmatrix}
  c \Delta t & \Delta x \\
  \end{pmatrix} \begin{pmatrix}
  1 & 0 \\
  0 & -1 \\
  \end{pmatrix} \begin{pmatrix}
  c \Delta t \\
  \Delta x \\
  \end{pmatrix} = \begin{pmatrix}
  c \Delta t' & \Delta x' \\
  \end{pmatrix} \begin{pmatrix}
  1 & 0 \\
  0 & -1 \\
  \end{pmatrix} \begin{pmatrix}
  c \Delta t' \\
  \Delta x' \\
  \end{pmatrix} \\
  &= \begin{pmatrix}
  c\Delta t & \Delta x \\
  \end{pmatrix} \begin{pmatrix}
  \gamma & - \frac{\gamma v}{c} \\
  - \frac{\gamma v}{c} & \gamma \\
  \end{pmatrix} \begin{pmatrix}
  1 & 0 \\
  0 & -1 \\
  \end{pmatrix} \begin{pmatrix}
  \gamma & - \frac{\gamma v}{c} \\
  - \frac{\gamma v}{c} & \gamma \\
  \end{pmatrix} \begin{pmatrix}
  c\Delta t \\
  \Delta x \\
  \end{pmatrix}
\end{align*}
So we must have:
\[
  \begin{pmatrix}
  \gamma & - \frac{\gamma v}{c} \\
  - \frac{\gamma v}{c} & \gamma \\
  \end{pmatrix} \begin{pmatrix}
  1 & 0 \\
  0 & -1 \\
  \end{pmatrix}\begin{pmatrix}
  \gamma & - \frac{\gamma v}{c} \\
  - \frac{\gamma v}{c} & \gamma \\
  \end{pmatrix} = \begin{pmatrix}
  1 & 0 \\
  0 & -1 \\
  \end{pmatrix}
\]
This Matrix is called the \textit{Minkowski} metric.
We saw this briefly at the end of IA Vectors and Matrices.
\begin{remark}
  We can compare this to how if we have a rotation matrix $R$, we have that $R R^{\trans} = R I R^{\trans} = I$, so they preserve the euclidean metric/norm.
\end{remark}
\begin{remark}[Warning]
  Unlike the euclidean metric, the Minkowski metric is \textbf{not} positive definite, that is, we can have $(\Delta s)^2 < 0$.
\end{remark}
\begin{definition}[Seperation of Events]
  If we have two events $P_1$ and $P_2$ then:
  \begin{enumerate}
    \item If $(\Delta s)^2 > 0$, then we say they are \textit{time like} separated.
    \item If $(\Delta s)^2 < 0$, then we say they are \textit{space like} separated.
    \item If $(\Delta s)^2 = 0$, then we say they are \textit{light like} separated.
  \end{enumerate}
\end{definition}
\begin{remark}
  If $(\Delta s)^2 = 0$, then $c^2(\Delta t)^2 = (\Delta x)^2$ so the two points must be connected by a light ray.
\end{remark}
Similarly to how the euclidean metric measures distances in space only, the Minkowski metric measures distances in space-time.
\subsection{Rapidity}
  To make the analogy with rotations clear, we can define rapidity.
  \begin{definition}[Rapidity]
    For a Lorentz transformation with factor $\gamma$, the \textit{rapidity} $\varphi$ is defined to be:
    \[
      \gamma \equiv \cosh \varphi
    \]
  \end{definition}
  We can then show that:
  \[
    \sinh \varphi = \sqrt{\cosh^2 \varphi - 1} = \sqrt{\gamma^2 -1 } = \frac{v \gamma}{c}
  \]
  So the Lorentz transformation matrix becomes:
  \[
    \begin{pmatrix}
    \gamma & - \frac{\gamma v}{c} \\
    - \frac{\gamma v}{c} & \gamma \\
    \end{pmatrix} = \begin{pmatrix}
    \cosh \varphi & -\sinh \varphi \\
    -\sinh \varphi &  \cosh \varphi \\
    \end{pmatrix} \equiv \Lambda(\varphi)
  \]
  If we do repeated Lorentz transformations, then we see that radpidities sum:
  \[
    \Lambda(\varphi_1)\Lambda(\varphi_2) = \Lambda(\varphi_1 + \varphi_2)
  \]
  This is similar to how composing 2D rotation sums angles.

  If we instead consider $\Lambda$ in terms of velocity, then we retrieve the formula for addition of velocities from \cref{additionOfVelocities}:
  \[
    \Lambda(v_1)\Lambda(v_2) = \Lambda\left(\frac{v_1 + v_2}{1 + \frac{v_1 v_2}{c^2}}\right)
  \]
\section{Lorentz Transformations in 4 Dimensions}
\subsection{Lorentz Groups}
We can generalise the Minkowski metric to 4 Dimensions (1 time dimension and 3 spatial dimensions) as:
\[
  \eta = \begin{pmatrix}
  1 & 0 & 0 & 0 \\
  0 & -1 & 0 & 0 \\
  0 & 0 & -1 & 0 \\
  0 & 0 & 0 & -1 \\
  \end{pmatrix}
\]
\begin{remark}[Notation]
  If we are using indices on this matrix, we typically use Greek letter $\mu$ and $\nu$ instead of $i$ and $j$, for example $\eta_{\mu \nu}$.
  Furthermore, by convention, we also take $\mu$ and $\nu$ running from 0 to 3 instead of 1 to 4 as we would typically expect.
\end{remark}

An event in space-time is given by a \textit{4-vector}:
\[
  X = (ct, x, y, z)
\]
\begin{remark}[Notation]
  \begin{itemize}
    \item By convention, we will use capital letters without boldface/underlines to denote a 4-vector.
    \item With indices, we again use Greek letters.
      We also use upper indices, for example $X^{\mu}$.
      Although not important here, in future courses, it will be important to distinguish \textit{upper and lower indices} for 4-vectors
  \end{itemize}
\end{remark}
The invariant distance between the origin and the event $X$ is given by the inner product:
\begin{equation}
  X \cdot X \equiv X^{\trans} \eta X = c^2 t^2 - x^2 - y^2 - z^2 \label{minkowskiDot}
\end{equation}
\begin{remark}
  In index notation this would be $X^{\mu}\eta_{\mu \nu}X^{\nu}$.
\end{remark}
As discussed earlier, this inner product is not positive definite:
\begin{center}
\begin{tabular}{c|c}
Condition & Name \\
\hline
$X \cdot X > 0$ & \textit{time-like} 4-vector  \\
$X \cdot X < 0$ & \textit{space-like} 4-vector \\
$X \cdot X = 0$ & \textit{light-like} or \textit{null} 4-vector \\
\end{tabular}
\end{center}
Four dimensional Lorentz transformations are $4 \times 4$ matrices $\Lambda$ such that:
\[
  X' = \Lambda X
\]
The defining feature of Lorentz transformations is that they leave this inner product invariant.
That is, $X' \cdot X' = X \cdot X$.
This happens if and only if $\Lambda^{\trans} \eta \Lambda = \eta$, i.e. they preserve the Minkowski metric.

Each $\Lambda$ has $4 \times 4 = 16$ components, both sides of $\Lambda^{\trans} \eta \Lambda = \eta$ are symmetric and so we have 10 constraints.
We therefore expect to find $6$ families of transformations from the unconstrained components:
\begin{itemize}
  \item There are \textbf{three families of euclidean rotations/reflections} in the form:
    \[
      \Lambda = \begin{pmatrix}
      1 & 0 & 0 & 0 \\
      0 &  &  &  \\
      0 &  & R &  \\
      0 &  &  &  \\
      \end{pmatrix}
    \]
    where $R^{\trans}R = \id_3$.
    Each family corresponds to each of the three spatial axes.
  \item There are \textbf{three families of Lorentz boosts}, one for each of the spatial axes:
    For example, along the $x$ and $y$ axes:
    \[
      \Lambda_x = \begin{pmatrix}
      \gamma & -\frac{\gamma v}{c} & 0 & 0 \\
      -\frac{\gamma v}{c} & \gamma & 0 & 0 \\
      0 & 0 & 1 & 0 \\
      0 & 0 & 0 & 1 \\
      \end{pmatrix}\quad
      \Lambda_y = \begin{pmatrix}
      \gamma & 0 & -\frac{\gamma v}{c} & 0 \\
      0 & 1 & 0 & 0 \\
      -\frac{\gamma v}{c} & 0 & \gamma & 0 \\
      0 & 0 & 0 & 1 \\
      \end{pmatrix}
    \]
    and similarly along other axes.
    We can then multiply these matrices to create a Lorentz boost along an arbitrary axis.
\end{itemize}
\begin{definition}[Lorentz Group]
  The set of all Lorentz transformations $\Lambda$ forms the \textit{Lorentz group}, denoted $\operatorname{O}(1, 3)$.
\end{definition}
\begin{remark}
  Because the inner product is preserved by this group, if a 4-vector is light-like in one frame, then it must also be light-like in any other frame and so it follows that the speed of light is the same in all frames.
\end{remark}
\begin{definition}[Proper Lorentz Group]
  The subgroup of the Lorentz group with $\det \Lambda = 1$ is called the \textit{proper Lorentz group}, denoted $\operatorname{SO}(1, 3)$.
\end{definition}
The proper Lorentz group removes reflections in either time or space, but not the combination of a reflection in time and space.
\begin{definition}[Proper Orthochronous]
  The subgroup of the proper Lorentz group that does not include those which reverse the time direction i.e. $\Lambda_{00} > 0$, is called the \textit{proper orthochronous} Lorentz group, denoted $\operatorname{SO}^{+}(1, 3)$.
\end{definition}
\begin{example}
  For example, the Lorentz transform:
  \[
    \Lambda = \begin{pmatrix}
    -1 & 0 & 0 & 0 \\
    0 & -1 & 0 & 0 \\
    0 & 0 & -1 & 0 \\
    0 & 0 & 0 & -1 \\
    \end{pmatrix}
  \]
  is in $\operatorname{SO}(1, 3)$ but not in $\SO^{+}(1, 3)$ because it reverses the time direction, despite the fact that $\det \Lambda = 1$.
\end{example}
\begin{remark}[Nomenclature]
  Typically when people refer to Lorentz transformations, they mean proper orthochronous Lorentz transformations, as we typically do not wan to consider reflections in space or time.
\end{remark}
\subsection{Proper Time}
We would like to define a new concept of velocity that is a 4-vector and so transforms nicely under Lorentz transforms.
To achieve this, we will first find a time $\tau$ that is \textit{invariant} under Lorentz transformations and define the \textit{4-velocity} as:
\[
  U \equiv \deriv{X}{\tau}
\]
\begin{remark}[Notation]
  We often abbreviate the spatial component of a 4-vector as $(ct, \vec{x})$ where $\vec{x} \in \R^3$.
\end{remark}
Suppose we have two points along a worldline $(ct_1, \vec{x}_1)$ and $(ct_2, \vec{x}_2)$.
We know that the invariant interval $(\Delta s)^2$ is the same in all inertial frames.
\begin{definition}[Proper Time]
  The \textit{proper time} between two points on a worldline is defined as:
  \[
    \Delta \tau \equiv \frac{\Delta s}{c}
  \]
\end{definition}
\begin{remark}[Note]
  Because we are considering two points on a worldline and worldlines must be contained within the light cone, the two events must be time-like, so $(\Delta s)^2 > 0$ and so $\Delta s \in \R^{+}$.
\end{remark}
Since all frames agree on $\Delta s$ and $c$, they also must agree on $\Delta \tau$.
We can then parameterise any worldline using $\tau$ as $\vec{x}(\tau)$ and $t(\tau)$.
Since all frames agree on $\tau$, they will also agree on this parameterisation.
\begin{remark}
  Proper time is like a Lorentzian version of the euclidean arc length along a curve.
  This is because in euclidean geometry, rotated frames might not agree on the positions of the endpoints of the curve, but will agree on the arc length between them.
  So in Lorentzian geometry, transformed frames might not agree on the space-times of a pair of events, but will agree on the $\Delta \tau$ between them in the parameterisation.
\end{remark}
Along a small segment of a parameterised worldline, we have:
\begin{align*}
  \d{\tau} &= \frac{1}{c}\sqrt{(\Delta s)^2} \\
           &= \sqrt{\d{t^2} - \frac{1}{c^2}(\d{x^2} + \d{y^2} + \d{z^2})} \\
           &= \d{t} \sqrt{1 - \frac{1}{c^2} \left[\left(\deriv{x}{t}\right)^2 + \left(\deriv{y}{t}\right)^2 + \left(\deriv{z}{t}\right)^2\right]} \\
           &= \d{t} \sqrt{1 - \frac{1}{c^2} \abs{\deriv{\vec{x}}{t}}^2}
\end{align*}
\begin{definition}[3-Velocity]
  We define the \textit{3-velocity} $\vec{u}$ as the standard instantaneous velocity of the spatial components, that is:
  \[
    \vec{u} = \deriv{\vec{x}}{t}
  \]
\end{definition}
We then have:
\[
  \d{\tau} = \d{t}\sqrt{1 - \frac{|\vec{u}|^2}{c^2}} = \frac{1}{\gamma}\d{t}
\]
where $\gamma$ is the Lorentz factor as a function of the instantaneous 3-velocity:
\[
  \gamma(u) = \frac{1}{\sqrt{1 - \frac{u^2}{c^2}}}
\]
Finally, we then have:
\[
  \deriv{t}{\tau} = \gamma
\]
Note that $\gamma$ does not have to be constant as $\vec{u}$ can be changing along the worldline.

We also conclude that a clock following the worldline has $\d{\vec{x}'} = 0 \implies \d{\tau} = \d{t'}$.
This means that the proper time is the time measured by an observer (not necessarily inertial) travelling along the world line.
\subsection{4-Velocity}
We now know that $\tau$ is invariant and we can parameterise a trajectory along a worldline as the 4-vector:
\[
  X(\tau) = \begin{pmatrix}
  ct(\tau) \\
  \vec{x}(\tau) \\
  \end{pmatrix}
\]
Since this is a 4-vector, we know that it transforms by $X' = \Lambda X$, so using that $\deriv{t}{\tau} = \gamma$, we have:
\[
  \deriv{X}{\tau} = \begin{pmatrix}
  c \deriv{t}{\tau} \\
  \deriv{\vec{x}}{\tau} \\
  \end{pmatrix} = \deriv{t}{\tau}\begin{pmatrix}
  c \\
  \deriv{\vec{x}}{t} \\
  \end{pmatrix} = \gamma \begin{pmatrix}
  c \\
  \vec{u} \\
  \end{pmatrix}
\]
\begin{definition}[4-Velocity]
  The \textit{4-velocity} $U$ is defined as:
  \[
    U = \deriv{X}{\tau} = \gamma \begin{pmatrix}
    c \\
    \vec{u} \\
    \end{pmatrix}
  \]
  where $\vec{u}$ is the three velocity and $\gamma$ is a function of $u = |\vec{u}|$.
\end{definition}
\begin{remark}
  For our purposes, a 4-vector will be any quantity $X$ that transforms as $X' = \Lambda X$.
\end{remark}
Since this is also a 4-vector, it transforms as $U' = \Lambda U$.
This relativistic 4-vector incorporates the familiar 3-velocity $\vec{u}$ into a nice Lorentzian object.

In particular, since $U' = \Lambda U$ and $\Lambda$ preserves the inner product, we have that $U' \cdot U' = U \cdot U$ in all frames.
If we compute $U \cdot U$, we see that it is just $c^2$:
\begin{equation}
  U \cdot U = \gamma^2 c^2 - \gamma^2 |\vec{u}|^2 = (c^2 - |\vec{u}|^2)\gamma^2 = c^2 \label{udotu}
\end{equation}
which is constant in all frames, as expected.
We can also use this to find a 4-velocity given 3-velocity $\vec{u}$ by solving $U \cdot U = c^2$ for the time component.
\subsection{4-Momentum}
Now we have defined a 4-velocity, we can define a \textit{4-momentum}:
\begin{definition}[4-Momentum]
  The \textit{4-momentum} $P$ of a particle is defined as:
  \[
    P \equiv mU = \begin{pmatrix}
    m\gamma c \\
    m\gamma \vec{u} \\
    \end{pmatrix}
  \]
  where $m$ is a property of the particle called the \textit{rest mass}.
\end{definition}
\begin{definition}[Relativistic Energy]
  The \textit{relativistic energy} $E$ is defined as:
  \[
    E \equiv m\gamma c^2
  \]
\end{definition}
\begin{definition}[Relativistic 3-Momentum]
  The \textit{relativistic 3-momentum} $\vec{p}$ is defined as:
  \[
    \vec{p} \equiv m\gamma \vec{u}
  \]
\end{definition}
We can then write the 4-vector $P$ as:
\[
  P = \begin{pmatrix}
  E/c \\
  \vec{p} \\
  \end{pmatrix}
\]
Because this is a 4-vector, by definition, it transforms as $P' = \Lambda P$.
The 4-momentum combines energy and momentum, analogously to how $X$ combines time and space.

In the absence of forces, we postulate that $P$ should be conserved, that is:
\begin{equation}
  \deriv{P}{\tau} = 0 \label{eqnPConserved}
\end{equation}
This is a relativistic generalisation of Newton's 2nd Law (\cref{newton2} -- in the case of $\vec{F} = \vec{0}$).
Because $P$ describes both the relativistic energy and the 3-momentum, \cref{eqnPConserved} combines both the conservation of energy and the conservation of momentum.
\begin{remark}[Note]
  The above equation is not something that we can derive, it is taken as an additional postulate.
\end{remark}

From \cref{udotu}, we know that $U \cdot U = c^2$ so:
\begin{equation}
  P \cdot P = (mU) \cdot (mU) = m^2 c^2 \label{pdotp}
\end{equation}
We can also calculate $P \cdot P$ in terms of the relativistic energy and 3-velocity using \cref{minkowskiDot}:
\begin{align}
  P \cdot P = \frac{E^2}{c^2} - |\vec{p}|^2 \implies E^2 = |\vec{p}|^2 c^2 + m^2 c^4 \label{4MomentumDot}
\end{align}
We can also check if the 4-momentum is consistent in the non-relativistic limit.
We defined $\vec{p} = m\gamma \vec{u}$ and in the non-relativistic limit $\frac{|\vec{u}|}{c^2} \ll 1 \implies \gamma \approx 1$, so we find that $\vec{p} \approx m \vec{u}$ which is the non-relativistic momentum.
We can expand the energy as:
\begin{align*}
  E = m\gamma c^2 &= \frac{mc^2}{\sqrt{1 - \frac{u^2}{c^2}}} \\
                  &= mc^2 \left(1 + \frac{1}{2}\cdot\frac{u^2}{c^2} + \cdots\right) \\
                  &= mc^2 + \frac{1}{2}mu^2 + \cdots
\end{align*}
In the non-relativistic limit, further terms in this series are negligible, so we obtain the non-relativistic kinetic energy $\frac{1}{2}mu^2$ but also an $mc^2$ term which is called the \textit{rest mass energy}.
This is an important consequence of relativity and suggests that mass leads to an energy.

We can also consider the high velocity limit as $|\vec{u}| \to c$.
By definition:
\[
  \vec{p}= \gamma m \vec{u}
\]
We sometimes call $\gamma m$ the \textit{relativistic mass}.
As $|\vec{u}| \to c$, $\gamma \to \infty$ so the relativistic mass $\gamma m \to \infty$.
This means that more and more force is needed to accelerate the particle as its speed increases.
In particular, any finite force cannot accelerate the particle to the speed of light.
A \textit{massive particle} (i.e. one with $m \neq 0$) must therefore have $|\vec{u}| < c$.
\subsection{Massless Particles}
In Galilean physics, the notion of a \textit{massless particle} does not make sense as the momentum of the particle would be identically 0.

Earlier, we derived the relativistic expression $P \cdot P = m^2 c^2$.
Therefore, taking $m = 0$ suggests that the 4-momentum of a massless particle should satisfy:
\[
  P \cdot P = 0
\]
This \textbf{does not} imply that $P = 0$ since the \textbf{Minkowski metric is not positive definite}, it instead tells us that the 4-momentum of a massless particle is a \textbf{light-like 4-vector} and so \textbf{lies along a light ray}.

Another property of massless particles is that their 4-velocity does not exist.
This is because if $P$ is finite and $m = 0$, we would need $\vec{u}$ to be infinite.
Instead, for a massless particle, the 4-momentum is the more ``fundamental'' concept.
Since $P \cdot P = 0$, from \cref{4MomentumDot}, we must have:
\[
  \frac{E^2}{c^2} = |\vec{p}|^2
\]
Thus the 4-momentum for a massless particle must be in the form:
\begin{equation}
  P = \frac{E}{c} \begin{pmatrix}
  1 \\
  \uvec{n} \\
  \end{pmatrix} \label{massless4Momentum}
\end{equation}
where $\uvec{n}$ is a unit 3-vector.

We can then interpret its 3-velocity as $\vec{u} \equiv \vec{n} c$ so since $\uvec{n} \cdot \uvec{n} = 1$, $|\vec{u}|^2 = c^2$.
This means that massless particles must move at the speed of light.

So for a massive particle $u \ll c$, and for a massless particle $u = c$.
Furthermore, along a light ray we have $\d{\tau} = 0$ because any two points on the light ray are light like separated so the invariant interval $(\Delta s)^2 = 0$.
This means that no time passes for a massless particle.

There are only two known massless particles: photons and gravitons.
In this course, we will only consider photons, which are ``particles of light''.
To proceed, we need to borrow from Quantum Mechanics that the energy of a photon is:
\[
  E = \hbar \omega
\]
where $\hbar \approx \qty{1.05e-34}{\joule\second}$ is the \textit{reduced Planck constant} and $\omega$ is the \textit{angular frequency of light}:
\[
  \omega = \frac{2\pi c}{\lambda}
\]
where $\lambda$ is the wavelength of the light.

Length contraction (\cref{lengthContraction}) tells us that different observers with relative velocities will see different wavelengths of light and hence will observe different energies of light.
This is consistent with that fact that $\frac{E}{c}$ is just the time component of the four momentum $P$ of the light so will be transformed appropriately by a Lorentz transform and so we expect different observes to measure different energies for the light.
\section{Particle Physics}
Particle accelerators, like those at CERN, operate by colliding particles at relativistic speeds
The proper mathematical framework for this is developed in \textit{Quantum Field Theory}, but we can learn a lot just from the conservation of the relativistic 4-momentum, in particular, the 4-momentum $P$ must be the same before and after a particle collision.

The two basic processes of particle physics are \textit{particle decay} and \textit{particle collisions}.
\subsection{Particle Decay}
Heavy particles are often unstable and decay into lighter particles.
For example, the \textit{Higgs Boson} decays in about \qty{e-22}{\second}.
Because such particles decay so quickly, we detect them using their \textit{decay products}, for example, the Higgs boson can $h$ decay into 2 photons.

In Large Hadron Collider at CERN, we have measured that:
\[
  m_hc^2 \approx \qty{125}{\giga\electronvolt}
\]
This means that the rest mass $m_h$ of the Higgs boson is about $10^{3}$ times heavier than the rest mass of an electron.

Since the 4-momentum is conserved, we must have:
\[
  P_{h} = P_{\gamma} + P_{\gamma'}
\]
where $P_{\gamma}$ and $P_{\gamma'}$ are the 4-momenta of the photons $\gamma$ and $\gamma'$.
This equation is currently Lorentz invariant (i.e it is equivalent in all frames).
However, to proceed, we need to decide a frame to work in.
There are two ``canonical'' options:
\begin{itemize}
  \item The \textit{lab frame}, in which the particles are moving.
  \item The \textit{centre of mass} or \textit{centre of momentum} frame, in which the total 3-momentum vanishes.
\end{itemize}
The centre of momentum frame is often more convenient than the lab frame when studying particle decays that start with one particle as we pick the rest frame of the unstable particle so its 4-momentum vanishes.
\subsubsection{In the Centre of Momentum Frame}
The Higgs boson is at rest so its 3-momentum is $\vec{0}$ and its energy is the rest mass energy $E = m_h c^2$.
Its 4-momentum $P_h$ is then:
\[
  P_h = \begin{pmatrix}
  m_h c \\
  \vec{0} \\
  \end{pmatrix}
\]
Since the photons are massless, from \cref{massless4Momentum}, we know that their 4-momenta are of the form:
\[
  P_{\gamma} = \frac{E_{\gamma}}{c}\begin{pmatrix}
  1 \\
  \uvec{n} \\
  \end{pmatrix} \text{ and }
  P_{\gamma'} = \frac{E_{\gamma'}}{c}\begin{pmatrix}
  1 \\
  \uvec{n}' \\
  \end{pmatrix}
\]
for unit vectors $\uvec{n}$ and $\uvec{n}'$.

By the conservation of 4-momentum:
\begin{align*}
  P_h = P_\gamma + P_{\gamma'} &\implies \begin{pmatrix}
  m_h c \\
  \vec{0} \\
  \end{pmatrix} = \frac{E_{\gamma}}{c}\begin{pmatrix}
  1 \\
  \uvec{n} \\
  \end{pmatrix} + \frac{E_{\gamma'}}{c}\begin{pmatrix}
  1 \\
  \uvec{n}' \\
  \end{pmatrix} \\
  &\implies m_h c = \frac{E_{\gamma}}{c} + \frac{E_{\gamma'}}{c} \text{ and } \vec{0} = \frac{E_{\gamma}}{c}\uvec{n} + \frac{E_{\gamma'}}{c}\uvec{n}'
\end{align*}
By the second equation, since $\uvec{n}$ and $\uvec{n}'$ are normal vectors, $E_{\gamma} = E_{-\gamma}$ and $\uvec{n} = -\uvec{n}'$, so the two photons emerge with opposite 3-momenta and equal energy.
\begin{center}
\begin{tikzpicture}[>=stealth]
  \fill (0, 0) circle (2pt) node[below right] {$h$} node[above left, scale=0.5] {Decay};
  \draw[->, rotate=30] (0.2, 0) -- (2, 0) node[midway, above left] {$\gamma$};
  \draw[->, rotate=30] (-0.2, 0) -- (-2, 0) node[midway, below right] {$\gamma'$};
\end{tikzpicture}
\end{center}
Then by the first equation, since $E_{\gamma} = E_{-\gamma}$, each photon carries away half the rest mass energy of the Higgs particle:
\[
  E_{\gamma} = E_{\gamma'} = \frac{m_hc^2}{2}
\]
So the rest mass energy of the Higgs has been converted to kinetic energy of the photons.
\begin{remark}[Summary]
  In the centre of momentum frame, we start with a Higgs particle at rest, and when it decays, we get two photons with opposite 3-momenta and equal energy, each carrying away half the rest mass energy of the Higgs particle.
\end{remark}
\subsubsection{In the Lab Frame}
If we consider the same process in the lab frame, the Higgs particle is not at rest:
\begin{center}
\begin{tikzpicture}[>=stealth]
  \draw[->, thick] (-2, 0) -- (0, 0) node[midway, above] {$h$};
  \draw[->, rotate=30, thick] (0, 0) -- (2, 0) node[midway, above left] {$\gamma$};
  \draw[->, rotate=-45, thick] (0, 0) -- (2, 0) node[midway, below left] {$\gamma'$};
  \draw[dashed] (0, 0) -- (1.7, 0);
  \draw (0, 0) ++ (0:0.7) arc (0:30:0.7) node[midway, right, scale=0.9] {$\theta$};
  \draw (0, 0) ++ (0:0.7) arc (0:-45:0.7) node[midway, right, scale=0.9] {$\alpha$};
  \fill (0, 0) circle (1.5pt) node[above left, scale=0.5] {Decay};
\end{tikzpicture}
\end{center}
We want to know the angle $\theta$ one of the photons makes with the path of the Higgs?
The important strategy here is to use conservation of 4-momentum, as we did above, but also \textit{invariants}.
Here, \textit{invariants} are the inner products of 4-vectors as we know that they are the same in all frames.

Starting from the conservation of 4-momentum:
\begin{align*}
  P_h - P_\gamma = P_{\gamma'} \implies P_{\gamma'} \cdot P_{\gamma'} &= (P_{h} - P_{\gamma}) \cdot (P_{h} - P_{\gamma}) \\
                                                                      &= P_h \cdot P_h - 2P_h \cdot P_\gamma + P_{\gamma} \cdot P_{\gamma} \\
                                                                      &= m^2_h c^2 - 2P_h \cdot P_{\gamma} + P_{\gamma} \cdot P_{\gamma}
\end{align*}
Both photons are massless and therefore light-like so in any frame $P_{\gamma'} \cdot P_{\gamma'} = P_{\gamma} \cdot P_{\gamma} = 0$.
Therefore:
\[
  2 P_h \cdot P_{\gamma} = m_h^2 c^2
\]
To compute this inner product, we can write $P_h$ and $P_\gamma$ in their most general forms:
\[
  P_{h} = \begin{pmatrix}
  E_h/c \\
  \vec{p}_{h} \\
  \end{pmatrix} \text{ and } P_{\gamma} = \frac{E\gamma}{c} \begin{pmatrix}
  1 \\
  \uvec{n} \\
  \end{pmatrix}
\]
for a unit vector $\uvec{n}$.
Substituting in, we then have:
\begin{align*}
  m_h^2 c^2 &= 2\left(\frac{E_h}{c} \cdot \frac{E_{\gamma}}{c} - \frac{E_{\gamma}}{c} \vec{p}_h \cdot \uvec{n}\right) \\
            &= 2 \left(\frac{E_h E_\gamma}{c^2} - \frac{E\gamma}{c}|\vec{p}_h|\cos\theta\right)
\end{align*}
We also know from \cref{4MomentumDot} that $E_h^2 = |\vec{p}_h|^2 c^2 + m_h^2 c^4$, so given $E_h$ and $E_\gamma$, we can then solve for the angle $\theta$.
\begin{remark}[Example Sheet]
  In Example Sheet 4 Q6, you will explore an example of a particle decaying into two massive particles instead of two massless particles.
\end{remark}
\subsection{Particle Collisions}
Another important particle interaction are particle collisions.
Consider two particles of mass $m$ that collide and then scatter.
\subsubsection{In the Centre of Momentum Frame}
Suppose that, in the centre of momentum frame, the particles scatter at an angle $\theta$ and the incoming particles have opposite 3-momenta of equal magnitude:
\begin{center}
\begin{tikzpicture}[>=stealth]
  \draw[->, thick] (-2, 0) -- (0, 0) node[midway, above] {1};
  \draw[->, thick] (2, 0) -- (0, 0) node[midway, below] {2};
  \draw[rotate=30, ->, dashed, thick] (0, 0) -- (2, 0) node[midway, above left] {3};
  \draw[rotate=30, ->, dashed, thick] (0, 0) -- (-2, 0) node[midway, below right] {4};
  \draw (0, 0) ++ (0:0.7) arc (0:30:0.7) node[midway, right, scale=0.9] {$\theta$};
\end{tikzpicture}
\end{center}
where particles 1 and 2 are the incoming particles and 3 and 4 are the outgoing particles.
The total 4-momentum before is $P_1 + P_2$ and after it is $P_3 + P_4$ so:
\begin{equation}
  P_1 + P_2 = P_3 + P_4 \tag{$\dag$} \label{eqnCollisionCoM}
\end{equation}
We know that $\vec{p}_1 + \vec{p}_2 = \vec{0}$ so equating 3-momentum components of \cref{eqnCollisionCoM}, we must also have $\vec{p}_3 + \vec{p}_4 = \vec{0}$.
We know that $|\vec{p}_1| = |\vec{p}_2|$ and now also that $|\vec{p}_3| = |\vec{p}_4|$.
From \cref{4MomentumDot}, we know that if $|\vec{p}_1| = |\vec{p}_2|$ and their masses are equal, their energies are also and so $E_1 = E_2$ and $E_3 = E_4$.
The time component of \cref{eqnCollisionCoM} then tells us that $E_1 = E_2 = E_3 = E_4$.
Thus:
\[
  |\vec{p}_1| = |\vec{p}_2| = |\vec{p}_3| = |\vec{p}_4|
\]
This means that any angle $\theta$ is allowed but momenta and energy of each particle must be the same before and after the collision.
All momenta are all in the same plane so, for example, we can pick:
\begin{align*}
  \vec{p}_1 &= |\vec{p}_1|(1, 0, 0)^{\trans} = - \vec{p}_2 \\
  \vec{p}_3 &= |\vec{p}_3|(\cos\theta, \sin\theta, 0)^{\trans} = - \vec{p}_4
\end{align*}
which satisfies the constraints.
\subsubsection{In the Lab Frame}
Now consider the same situation in the lab frame where we pick a frame attached to $P_2$.
In this frame, $P_2$ is at rest and is hit by $P_1$.
We want to find the angle $\phi$:
\begin{center}
\begin{tikzpicture}[>=stealth]
  \draw[->, thick] (-2, 0) -- (0, 0) node[midway, above] {$P_1$};
\fill (0, 0) circle (2pt) node[below] {$P_2$};
  \draw[->, rotate=30, thick] (0, 0) -- (2, 0) node[midway, above left] {$P_3$};
  \draw[->, rotate=-45, thick] (0, 0) -- (2, 0) node[midway, below left] {$P_4$};
  \draw[dashed] (0, 0) -- (1.7, 0);
  \draw (0, 0) ++ (0:0.7) arc (0:30:0.7) node[midway, right, scale=0.9] {$\phi$};
  \draw (0, 0) ++ (0:0.7) arc (0:-45:0.7) node[midway, right, scale=0.9] {$\alpha$};
\end{tikzpicture}
\end{center}
We can again use invariants.
Since we do not care about what $P_4$ is doing, we can rearrange so that $P_4$ is on its own side so that when we take the inner product of both sides with themselves, all the information about $P_4$ is removed since $P_4 \cdot P_4 = m^2 c^2$ from \cref{pdotp}.
\begin{align}
  P_1 + P_2 = P_3 + P_4 \implies (P_1 + P_2 - P_3) \cdot (P_1 + P_2 - P_3) &= P_4 \cdot P_4 \nonumber \\
  \implies P_1 \cdot P_1 + P_2 \cdot P_2 + P_3 \cdot P_3 - 2P_1 \cdot P_3 - 2 P_2 \cdot P_3 + 2 P_1 \cdot P_2 &= m^2 c^2 \nonumber \\
  3m^2 c^2 - 2P_1 \cdot P_3 - 2 P_2 \cdot P_3 + 2 P_1 \cdot P_2 &= m^2 c^2 \label{eqnCollisionLab} \tag{$\star$}
\end{align}
To take these inner products, we need to determine the form of $P_1$, $P_2$ and $P_3$:
We don't know anything about $P_1$ and $P_3$ but we do know that $P_2$ is at rest so has 3-momentum $\vec{0}$ and has rest mass energy:
\[
  P_2 = \begin{pmatrix}
  mc \\
  \vec{0} \\
  \end{pmatrix},\ P_1 = \begin{pmatrix}
  E_1/c \\
  \vec{p}_1 \\
  \end{pmatrix},\ P_3 = \begin{pmatrix}
  E_3/c \\
  \vec{p}_3 \\
  \end{pmatrix}
\]
Therefore:
\[
  -2P_2 \cdot P_3 + 2P_1 \cdot P_2 = -2E_3m + 2 E_1 m
\]
and:
\[
  -2P_1 \cdot P_3 = -2 \left(\frac{E_1E_3}{c^2} - \vec{p}_1 \cdot \vec{p}_3\right) \\
\]
Substituting back into \cref{eqnCollisionLab}, we have:
\[
  2m^2 c^2 + 2m(E_1 - E_3) - \frac{2E_1E_3}{c^2} = -2|\vec{p}_1||\vec{p}_3|\cos\phi
\]
So if we are given $E_1$ and $E_3$, similarly to in the lab frame for particle decay, we can use $E_i^2 = |\vec{p}_i|^2 c^2 + m^2 c^4$ to eliminate the $|\vec{p}_1|$ and $|\vec{p}_3|$ and therefore obtain $\phi$ in terms of $E_1$ and $E_3$.
\begin{remark}[Example Sheet]
  In Example Sheet 4 Q7, you will explore an example of \textit{Compton scattering}.
  This is when a photon scatters off a massive particle.
\end{remark}
\end{document}
