\documentclass[../main.tex]{subfiles}
\begin{document}
\chapter{Linear Programming}
\section{Formulation}
\begin{definition}[Linear Program]
  A \textit{linear program} has general form:
  \begin{mini*}
    {\vec{x}}{\vec{c}^{\trans}\vec{x}}{}{\label{linearGeneral}}
    \addConstraint{\vec{a}^{\trans}_i \vec{x}}{\geq b_i,\ i \in M_1}
    \addConstraint{\vec{a}^{\trans}_i \vec{x}}{\leq b_i,\ i \in M_2}
    \addConstraint{\vec{a}^{\trans}_i \vec{x}}{= b_i,\ i \in M_3}
    \addConstraint{x_j}{\geq 0,\ j \in N_1}
    \addConstraint{x_j}{\leq 0,\ j \in N_2}
    \addConstraint{x_j}{\text{ free},\ j \in N_3}
  \end{mini*}
\end{definition}
We can in fact write such constraints as a matrix:
\[
  A = \begin{pmatrix}
  \leftarrow \vec{a}^{\trans}_1 \rightarrow \\
  \leftarrow \vec{a}^{\trans}_2 \rightarrow \\
  \cdots \\
  \leftarrow \vec{a}^{\trans}_m \rightarrow
  \end{pmatrix}
\]
\begin{theorem}
  The dual of the problem \cref{linearGeneral} is given by:
\end{theorem}
\begin{proof}
  Adding slack variables, we have:
  \begin{align*}
    L(\vec{x}, \vec{s}, \vec{\lambda}) &= \vec{c}^{\trans} \vec{x} - \sum_{i \in M_1} \lambda_i (\vec{a}^{\trans}_i \vec{x} - s_i - b_i) \\
                     &+ \sum_{i \in M_2} \lambda_i(\vec{a}_i^{\trans} \vec{x} + s_i - b_i) - \sum_{i \in M_3} \lambda_i(\vec{a}^{\trans}_i \vec{x} - b_i) \\
                     &= \sum_{j \in N_1} (c_j - \vec{\lambda}^{\trans} A_j)\vec{x}_j + \sum_{j \in N_2} (c_j - \vec{\lambda}^{\trans} A_j)\vec{x}_j + \sum_{j \in N_3} (c_j - \vec{\lambda}^{\trans} A_j)\vec{x}_j \\
                     &+\sum_{i \in M_1} \lambda_i s_i - \sum_{i \in M_2} \lambda_i s_i + \sum_{i \in M_1 \cup M_2 \cup M_3} \lambda_i b_i
  \end{align*}
  where $A_i$ is the $i$-Th column of $A$.
  Consider the set:
  \[
    \Lambda = \left\{\vec{\lambda}: \begin{cases}
    \lambda_i \geq 0 & \text{ for } i \in M_1 \\
    \lambda_i \leq 0 & \text{ for } i \in M_2 \\
    \lambda_i \text{ free} & \text{ for } i \in M_3 \\
    \vec{\lambda}^{\trans} A_j \leq c_j & \text{ for } j \in N_1 \\
    \vec{\lambda}^{\trans} A_j \geq c_j & \text{ for } j \in N_2 \\
    \vec{\lambda}^{\trans} A_j = c_j & \text{ for } j \in N_3
    \end{cases}\right\}
  \]
  If $\vec{\lambda} \in \Lambda$, then:
  \[
    \min_{\vec{x}, \vec{s} \geq \vec{0}} L(\vec{x}, \vec{s}, \vec{\lambda}) = \vec{b}^{\trans} \vec{\lambda} = g(\vec{\lambda})
  \]
  So we see that the dual is also a linear program.
\end{proof}
\section{Optimality Conditions}
\begin{theorem}
  Let $\vec{x}$ and $\vec{\lambda}$ be feasible solutions to the primal program \cref{linearGeneral}  and its respective $\vec{\lambda}$.
  We then have that these are optimal if and only if:
  \[
    \lambda_i (\vec{a}^{\trans}_i  \vec{x} - b_i) = 0\ \forall i \text{ and } (c_j - \vec{\lambda}^{\trans}A_j)x_j = 0\ \forall j
  \]
\end{theorem}
\begin{proof}
  Define $u_i = \lambda_i(\vec{a}^{\trans}_i \vec{x} - b_i)$ and $v_j = (c_j - \vec{\lambda}^{\trans} A_j)x_j$.
  Since $\vec{x}$ and $\vec{\lambda}$ are primal and dual feasible respectively, we have that $u_i \geq 0\ \forall i$ and $v_j \geq 0\ \forall j$.

  Taking the sum over all $i$ and $j$ respectively:
  \begin{align*}
    \sum_{i} u_i = \vec{\lambda}^{\trans} A \vec{x} - \vec{\lambda}^{\trans} \vec{b} \\
    \sum_{i} v_i = \vec{c}^{\trans} \vec{x} - \vec{\lambda}^{\trans} A\vec{x}
  \end{align*}
  and so:
  \[
    \sum_{i} u_i + \sum_{j} v_j = \vec{c}^{\trans} \vec{x} - \vec{\lambda}^{\trans} \vec{b}
  \]
  By weak duality, we know that $\vec{c}^{\trans} \vec{x} \geq \vec{\lambda}^{\trans} \vec{b}$.
  \begin{proofdirection}{Assume they are optimal}
    So we then have that $\vec{c}^{\trans} \vec{x} = \vec{\lambda}^{\trans} \vec{b}$, we must then have $u_i = 0\ \forall i$ and $v_j = 0\ \forall j$ as this is a sum of non-negative numbers equal to $0$.
  \end{proofdirection}
  \begin{proofdirection}{Assume the equality holds}
    It then follows that $\vec{c}^{\trans} \vec{x} = \vec{\lambda}^{\trans} \vec{b}$ and so $\vec{x}$ and $\vec{\lambda}$ are optimal.
  \end{proofdirection}
\end{proof}
\section{Standard Form for Linear Programs}
Observe that we can always write any linear program as:
\begin{mini*}
  {\vec{x}}{\vec{c}T \vec{x}}{}{}
  \addConstraint{A\vec{x}}{\leq \vec{b}}
\end{mini*}
This is the \textit{general form} for an LP.
We say that an LP is in \textit{standard form} if it is in the form:
\begin{mini*}
  {\vec{x}}{\vec{c}^{\trans} \vec{x}}{}{}
  \addConstraint{A\vec{x}}{= \vec{b}}
  \addConstraint{\vec{x}}{\geq \vec{0}}
\end{mini*}
We then observe that any $x \in \R$ and be expressed as $x = x_{+} - x_{-}$ where $x_{+}, x_{-} \in \R^{+}$.
This can be used to convert any general form problem to a standard form problem.

Starting with:
\begin{mini*}
  {\vec{x}}{\vec{c}T \vec{x}}{}{}
  \addConstraint{A\vec{x}}{\leq \vec{b}}
\end{mini*}
we first add slack variables:
\begin{mini*}
  {\vec{x}}{\vec{c}^{\trans} \vec{x}}{}{}
  \addConstraint{A\vec{x} + \vec{s}}{= \vec{b}}
  \addConstraint{\vec{s}}{\geq 0}
\end{mini*}
and can then use the trick:
\begin{mini*}
  {\vec{x}}{\vec{c}^{\trans}(\vec{x}_{+} - \vec{x}_{-})}{}{}
  \addConstraint{A(\vec{x}_{+} - \vec{x}_{-}) + \vec{s}}{= \vec{b}}
  \addConstraint{\vec{x}_{+}}{\geq 0}
  \addConstraint{\vec{x}_{-}}{\geq 0}
  \addConstraint{\vec{s}}{\geq 0}
\end{mini*}
which is now a standard form problem in $(\vec{x}_{+}, \vec{x}_{-}, \vec{s})$.
\section{Solving Linear Programs}
We first note that minimising $\vec{c}^{\trans} \vec{x}$ is equivalent to maximising $-\vec{c}^{\trans} \vec{x}$.
But $-\vec{c}^{\trans} \vec{x}$ is a convex function, so we are essentially trying to maximise a convex function over some restricted region.
\begin{definition}[Extreme Point]
  Let $C \subseteq \R^{n}$ be a convex set.
  A point $\vec{x} \in C$ is called an \textit{extreme point} if it cannot be written as a convex combination of two distinct points in $C$.

  That is, for all $\vec{y}, \vec{z} \in \C$ and $\delta \in (0, 1)$:
  \[
    \vec{x} = \delta \vec{y} + (1 - \delta) \vec{z} \implies \vec{x} = \vec{y} = \vec{z}
  \]
\end{definition}
\begin{remark}[Intuition]
  A point is extreme if it does not lie on any line segment \textbf{contained within} $C$.

  So if $C$ is a triangle, only its vertices are extreme points, and if $C$ is a circle, then only its boundary points are extreme points.
\end{remark}
\begin{lemma}
  The max of a convex function $f$ on a convex set $C$ occurs at an extreme point of $C$.
\end{lemma}
\begin{proof}
  Suppose $\vec{x}$ is not an extreme point.
  Thus there is $\vec{y}, \vec{z} \in C$ and $\delta \in (0, 1)$ such that:
  \[
    f(\vec{x}) \leq \delta f(\vec{y}) + (1 - \delta)f(\vec{z}) \leq \max\{f(\vec{y}), f(\vec{z})\}
  \]
\end{proof}
In a linear program, the constraint set is always a polytope.
Such regions always have finitely many extreme points and so if we can find a list of these extreme points it becomes easier to maximise $-\vec{c}^{\trans} \vec{x}$.
\subsection{Basic Solutions and Basic Feasible Solutions}
Recall that we wish to solve:
\begin{mini*}
  {\vec{x}}{\vec{c}^{\trans} \vec{x}}{}{}
  \addConstraint{A\vec{x}}{= \vec{b}}
  \addConstraint{\vec{x}}{\geq \vec{0}}
\end{mini*}
where:
\[
  A = \begin{pmatrix}
  \leftarrow \vec{a}^{\trans}_1 \rightarrow \\
  \leftarrow \vec{a}^{\trans}_2 \rightarrow \\
  \cdots \\
  \leftarrow \vec{a}^{\trans}_m \rightarrow
  \end{pmatrix}
\]
We will make three assumptions:
\begin{enumerate}
  \item The rows of $A$ are linearly independent.
    This can in fact be made without loss in generality as if two rows are not linearly independent then we have a redundant constraint so we can simply remove one of the rows.
  \item Every set of $m$ columns of $A$ is linearly independent.
    This cannot be made without loss in generality, but for most applications, it will typically hold.
\end{enumerate}
For our final assumption, we will need some additional definitions.
\begin{definition}[Basic Solution]
  We call a point $\vec{x}$ that satisfies $A\vec{x} = \vec{b}$ and has at most $m$ non-zero entries a \textit{basic solution}.
\end{definition}
\begin{definition}[Basic Feasible Solution]
  We call a basic solution $\vec{x}$ that satisfies $\vec{x} \geq \vec{0}$ a \textit{basic feasible solution}.
\end{definition}
Suppose we choose indices $\{B(1), \ldots, B(m)\}$ such that $x_i = 0$ for all $i \notin \{B(1), \ldots, B(m)\}$, i.e. we have a set of indices we know have $x_i = 0$ and a set of indices we know have $x_i \neq 0$:
\[
  A \vec{x} = \sum_{i = 1}^{n} \vec{A}_i x_i = \sum_{i = 1}^{m} \vec{A}_{B(i)}\vec{x}_{B(i)} = B \vec{x}_{B}
\]
where $\vec{A}_i$ is the $i$-th column of $A$ and:
\[
  B = \begin{pmatrix}
  \uparrow & & \uparrow  \\
  \vec{A}_{B(1)} & \cdots & \vec{A}_{B(m)}  \\
  \downarrow &  & \downarrow \\
  \end{pmatrix}
 \text{ and } \vec{x}_{B} = \begin{pmatrix}
  x_{B(1)} \\
  \vdots \\
  x_{B(m)} \\
  \end{pmatrix}
\]
Ideally for this problem, we want $A\vec{x} = \vec{b} \iff B\vec{x}_{B} = \vec{b} \iff \vec{x}_{B} = B^{-1} \vec{b}$.
\begin{definition}[Basic Variables, Matrix, and Columns]
  We call:
  \begin{itemize}
    \item $\vec{x}_{B(1)}, \ldots, \vec{x}_{B(m)}$ the \textit{basic variables},
    \item $B$ the \textit{basis matrix},
    \item $\vec{A}_{B(1)}, \ldots, \vec{A}_{B(m)}$ the \textit{basic columns}.
  \end{itemize}
\end{definition}
For every choice of $B$ (of which there are $\binom{n}{m}$ such choices for the columns), we obtain a basic solution.
So if $B^{-1}\vec{b} \geq 0$, then that the basis matrix $B$ leads to a basic feasible solution.

We can now define our third assumption:
\begin{enumerate}
  \setcounter{enumi}{2}
  \item Every basic feasible solutions is \textit{non-degenerate}.
    That is, it has \textbf{exactly} $m$ non-zero entries:
    \[
      B^{-1} \vec{b} \geq 0 \implies B^{-1} \vec{b} > 0
    \]
    This also cannot be made without loss in generality but is required to ensure that algorithms later in this chapter do not get stuck.
\end{enumerate}
\begin{theorem}
  A vector $\vec{x}$ is a basic feasible solution if and only if it is an extreme point of the set:
  \[
    \mathcal{X}(\vec{b}) = \{\vec{x}: A \vec{x} = \vec{b}, \vec{x} \geq 0\}
  \]
\end{theorem}
\begin{remark}[Rudimentary Algorithm for Solving a L.P.]
  This theorem means that we can solve an L.P. by just checking each of the basic feasible solutions, that is:
  \begin{itemize}
    \item List all $\binom{n}{m}$ basic solutions.
    \item Filter these basic solutions to just the basic feasible solutions.
    \item Evaluate $\vec{c}^{\trans} \vec{x}$ for each of these basic feasible solutions and choose the best one.
  \end{itemize}
\end{remark}
\begin{proof}
  \begin{proofdirection}{Assume $\vec{x}$ is a basic feasible solution}
    Suppose that $\vec{x}$ can be written as:
    \[
      \vec{x} = (1 - \delta)\vec{y} + \delta \vec{z}
    \]
    for some $\vec{y}, \vec{z} \in \mathcal{X}(\vec{b})$ and $\delta \in (0, 1)$.

    If $x_i = 0$, then:
    \[
      x_i = 0 = (1 - \delta)y_i + \delta z_i
    \]
    but we also have that $y_i, z_i \geq 0$ as $\vec{y}, \vec{z} \in \mathcal{X}(\vec{b})$ and so $y_i = z_i = 0$.
    Since $\vec{x}$ is a basic feasible solution, $\vec{y}$ and $\vec{z}$ must be basic solutions corresponding to the same basis as $\vec{x}$.
    Thus $\vec{x} = \vec{y} = \vec{z} = B^{-1} \vec{b}$ and so $\vec{x}$ is an extreme point of $\mathcal{X}(\vec{b})$.
  \end{proofdirection}
  \begin{proofdirection}{Assume $\vec{x}$ is not a basic feasible solution}
    We wish to show that $\vec{x}$ is not an extreme point of $\mathcal{X}(\vec{b})$.
    If $\vec{x} \notin \mathcal{X}(\vec{b})$, then obviously it is not an extreme point  of $\mathcal{X}(\vec{b})$ and so we assume that $\vec{x} \in \mathcal{X}(\vec{b})$.

    Since $\vec{x}$ is not a basic feasible solution, it must have $r > m$ positive entries.
    Suppose that $x_{i_1} > 0, x_{i_2} > 0, \ldots, x_{i_r} > 0$.
    Because we have chosen more columns than the rank $m$, the set of columns $\{\vec{A}_{i_1}, \ldots, \vec{A}_{i_r}\}$ must be linearly dependent.
    By definition this means that $\exists w_{i_1}, \ldots, w_{i_r}$ not all 0 such that:
    \[
      w_{i_1} \vec{A}_{i_1} + \cdots + w_{i_r} \vec{A}_{i_r} = \vec{0}
    \]
    Construct a vector $\vec{w}$ such that $\vec{w}$ is $0$ everywhere except that $i_1, \ldots, i_r$ where $w(i_1) = w_{i_1}, \ldots, w(i_r) = w_{i_r}$.

    Now observe that $A\vec{w} = \vec{0}$ because:
    \[
      A \vec{w} = \vec{A}_{i_1} w_{i_1} + \cdots + \vec{A}_{i_r} w_{i_r}
    \]
    Now consider two new points:
    \begin{align*}
      \vec{y} = \vec{x} + \varepsilon \vec{w} \\
      \vec{z} = \vec{x} - \varepsilon \vec{w}
    \end{align*}
    for some small $\varepsilon > 0$.

    For small enough $\varepsilon$, $\vec{y}, \vec{z} \in \mathcal{X}(\vec{b})$.
    Thus we can write $\vec{x}$ as:
    \[
      \vec{x} = \frac{1}{2}\vec{y} + \frac{1}{2}\vec{z}
    \]
    for $\vec{y}, \vec{z} \in \mathcal{X}(\vec{b})$ and so $\vec{x}$ cannot be an extreme point of $\mathcal{X}(\vec{b})$.
  \end{proofdirection}
\end{proof}
\subsection{Towards the Simplex Method}
When the L.P. is in standard form, the optimality conditions are:
\begin{enumerate}
  \item \textbf{Primal feasibility --} $A\vec{x} = \vec{b}$, $\vec{x} \geq 0$.
  \item \textbf{Dual feasibility --} $A^{\trans} \vec{\lambda} \leq \vec{c}$
  \item \textbf{Complementary Slackness --} $\vec{x}^{\trans}(\vec{c} - A^{\trans} \vec{\lambda}) = 0$
\end{enumerate}
For any basic feasible solution, the complementary slackness equation reduces to:
\[
  \vec{x}_{B}^{\trans}(\vec{c}_{B} - B^{\trans} \vec{\lambda}) = 0
\]
Since $\vec{x}_{B} > \vec{0}$ by the non-degeneracy assumption, we must have:
\[
  \vec{c}_{B} = B^{T} \vec{\lambda} \implies \vec{\lambda} = (B^{\trans})^{-1} C_{B}
\]
If $A^{\trans}(B^{\trans})^{-1}\vec{c}_{B} \leq \vec{c}$, then $\vec{\lambda}$ is dual feasible and so $\vec{x}$ is optimal.

So if we check $A^{\trans}(B^{\trans})^{-1}\vec{c}_{B} \leq \vec{c}$ at every basic feasible solution, any solution satisfying this must be optimal.
\end{document}
